---
title: "acute aggregate analysis"
output: html_document
date: "2026-03-25"
---

##note - not presenting these figures in the paper; rather, presenting the acute result with background exposures because the large difference in background concentrations at commons for some species, like benzene, is not an issue for the acute exposures##


#load libraries 
```{r}
library("dplyr")
library("tidyr")
library("Rmisc")
library("terra")
library("readxl")
library("tidyverse")
library("stringr")
library("lubridate")
library("ggh4x")
detach(package:Rmisc)
detach(package:plyr)
```



#load chronic "background" data
```{r}
file1 <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/chronic_spatial_trend_monthly_gapfilled.xlsx"

sheets <- excel_sheets(file1)

chronic_list <- lapply(sheets, function(s) {

  df <- read_excel(file1, sheet = s)

  df_long <- df %>%
    pivot_longer(
      cols = starts_with("Distance_"),
      names_to = "distance_ft",
      values_to = "concentration_ppb"
    ) %>%
    mutate(
      distance_ft = as.numeric(str_extract(distance_ft, "\\d+")),
      chemical = s
    )

  return(df_long)
})

chronic <- bind_rows(chronic_list)

chronic <- chronic %>%
  mutate(chemical = recode(
    chemical,
    "aromatic medium" = "aromatic_medium",
    "aliphatic low"   = "aliphatic_low",
    "nonane" = "n_nonane",
    "decane" = "n_decane"
  ))
```



#load acute trigger cannister data
```{r}
file2 <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/Broomfield-Triggered Canisters-VOC concentration vs Distances.xlsx"

acute_raw <- read_excel(file2)

meta_cols <- c(
  "Start Date Time",
  "Stop Date Time",
  "Source Wellpad",
  "Operation",
  "Sampling Site",
  "Distance (site, pad) ft",
  "Note"
)

acute <- acute_raw %>%
  select(
    all_of(meta_cols),
    matches("\\d+ft ppb$")   # keeps things like "benzene 500ft ppb"
  )

# reshape to long format
acute_long <- acute %>%
  pivot_longer(
    cols = -all_of(meta_cols),
    names_to = "chem_dist",
    values_to = "concentration_ppb"
  )

#separate chemical and distance
acute_long <- acute_long %>%
  mutate(
    chemical = str_extract(chem_dist, "^[^ ]+"),
    distance_ft = as.numeric(str_extract(chem_dist, "\\d+(?=ft)"))
  ) %>%
  select(-chem_dist)
```



#filter dataset and combine chemicals into groups 
```{r}
group_vars <- c(
  "Start Date Time",
  "Stop Date Time",
  "Source Wellpad",
  "Operation",
  "Sampling Site",
  "Distance (site, pad) ft",
  "distance_ft"
)

acute_long <- acute_long %>%
  rename(concentration = concentration_ppb)

#group chemical categories
## xylenes
xylenes_df <- acute_long %>%
  filter(chemical %in% c("o_xylene", "m_p_xylene")) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(concentration = sum(concentration, na.rm = TRUE)) %>%
  mutate(chemical = "xylenes")

## aliphatic_low
aliphatic_low_chems <- c(
  "i_pentane", "n_pentane", "n_hexane", "n_heptane", "n_octane",
  "cyclopentane", "cyclohexane", "methylcyclohexane",
  "2_2_4_trimethylpentane", "2_3_4_trimethylpentane",
  "2_4_dimethylpentane", "2_3_dimethylpentane",
  "2_methylhexane", "3_methylhexane",
  "2_methylheptane", "3_methylheptane",
  "t_2_pentene", "1_pentene", "c_2_pentene",
  "isoprene"
)

aliphatic_low_df <- acute_long %>%
  filter(chemical %in% aliphatic_low_chems) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(concentration = sum(concentration, na.rm = TRUE)) %>%
  mutate(chemical = "aliphatic_low")

## aromatic_medium
aromatic_medium_chems <- c(
  "i_propylbenzene", "n_propylbenzene",
  "2_ethyltoluene", "3_ethyltoluene", "4_ethyltoluene",
  "1_2_3_trimethylbenzene", "1_2_4_trimethylbenzene", "1_3_5_trimethylbenzene",
  "1_3_diethylbenzene", "1_4_diethylbenzene"
)

aromatic_medium_df <- acute_long %>%
  filter(chemical %in% aromatic_medium_chems) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(concentration = sum(concentration, na.rm = TRUE)) %>%
  mutate(chemical = "aromatic_medium")

#remove unwanted chems
drop_chems <- c(
  # components now grouped
  "o_xylene", "m_p_xylene",
  aliphatic_low_chems,
  aromatic_medium_chems,

  # explicitly remove
  "methane", "ethane", "propane",
  "i_butane", "n_butane",
  "ethene", "ethyne",
  "t_2_butene", "1_butene", "c_2_butene",
  "NMVOC"
)

acute_filtered <- acute_long %>%
  filter(!chemical %in% drop_chems)

#final clean dataset
acute_clean <- bind_rows(
  acute_filtered,
  xylenes_df,
  aliphatic_low_df,
  aromatic_medium_df
)

#checks
# check new chemical list
sort(unique(acute_clean$chemical))

# make sure grouped ones exist
table(acute_clean$chemical %in% c("xylenes", "aliphatic_low", "aromatic_medium"))

# check no duplicates per group
dup_check <- acute_clean %>%
  group_by(across(all_of(group_vars)), chemical) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1)

nrow(dup_check)

setdiff(group_vars, names(acute_clean))

setdiff(aliphatic_low_chems, unique(acute_long$chemical))
```


#calculate actual, 50th, and 95th percentile ozone concentrations
```{r}
ozone = read.csv("~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/ozone_hourly_2018-2023.csv")

#prepare data with time stamps
ozone_clean <- ozone %>%
  mutate(
    Date = as.Date(Date.Local),
    Hour = format(as.POSIXct(Time.Local, format = "%H:%M"), "%H"),
    ozone_ppb = Sample.Measurement * 1000   # ppm → ppb
  ) %>%
  filter(lubridate::year(Date) >= 2019 & lubridate::year(Date) <= 2022)

acute_time_index <- acute_clean %>%
  mutate(
    Date = as.Date(`Start Date Time`),
    Hour = format(as.POSIXct(`Start Date Time`), "%H")
  ) %>%
  distinct(Date, Hour)

#time-matched ozone concentrations (scenario 1)
ozone_time_matched <- ozone_clean %>%
  semi_join(acute_time_index, by = c("Date", "Hour")) %>%
  group_by(Date, Hour) %>%
  summarise(
    ozone_time_matched = median(ozone_ppb, na.rm = TRUE),
    .groups = "drop"
  )

#50th and 95th percentile concentrations
ozone_p50 <- median(ozone_clean$ozone_ppb, na.rm = TRUE)
ozone_p95 <- quantile(ozone_clean$ozone_ppb, 0.95, na.rm = TRUE)

#final ozone in one df
ozone_final <- ozone_time_matched %>%
  mutate(
    ozone_p50 = ozone_p50,
    ozone_p95 = ozone_p95
  )

#checks
# should match acute timestamps
nrow(ozone_final)
nrow(acute_time_index)

# check missing ozone values
sum(is.na(ozone_final$ozone_time_matched))

```


#match chemicl names to rfcs 
```{r}
rfcs = read.csv("~/CO_CRA/CO_CRA_aggregate/data/RfCs_5-2026.csv")

rfcs_acute <- rfcs %>%
  dplyr::filter(Exposure == "acute")

rfcs_acute$Chemical[rfcs_acute$Chemical == "Benzene"] <- "benzene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Toluene"] <- "toluene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Ethylbenzene"] <- "ethylbenzene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Xylenes"] <- "xylenes"
rfcs_acute$Chemical[rfcs_acute$Chemical == "n-nonane"] <- "n_nonane"
rfcs_acute$Chemical[rfcs_acute$Chemical == "n-decane"] <- "n_decane"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Aliphatic low"] <- "aliphatic_low"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Aromatic medium"] <- "aromatic_medium"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Styrene"] <- "styrene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Trichloroethylene (TCE)"] <- "c2hcl3"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Tetrachloroethylene (PCE)"] <- "c2cl4"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Propene"] <- "propene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Ozone"] <- "ozone"

#check
setdiff(acute_clean$chemical, rfcs_acute$Chemical)
setdiff(rfcs_acute$Chemical, acute_clean$chemical)
```


#calculate HQs
```{r}
# --- 1. Prepare chronic dataset to match acute ---
chronic_ready <- chronic %>%
  rename(concentration = concentration_ppb) %>%
  mutate(
    Operation = NA_character_,
    distance_ft = as.numeric(distance_ft)
  ) %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    chemical,
    concentration,
    Operation
  )

# --- 2. Combine acute + chronic concentrations ---
combined_df <- bind_rows(
  acute_clean %>%
    select(
      `Start Date Time`,
      `Source Wellpad`,
      `Sampling Site`,
      distance_ft,
      chemical,
      concentration,
      Operation
    ),
  chronic_ready
)

# Sum concentrations per timestamp, site, chemical, distance
combined_df <- combined_df %>%
  group_by(`Start Date Time`, `Source Wellpad`, `Sampling Site`, distance_ft, chemical) %>%
  summarise(
    concentration_total = sum(concentration, na.rm = TRUE),
    Operation = dplyr::first(na.omit(Operation)), # take acute Operation if available
    .groups = "drop"
  )

# --- 3. Calculate chemical HQs ---
chemical_hq <- combined_df %>%
  filter(chemical != "ozone") %>%
  left_join(
    rfcs_acute %>% select(Chemical, Endpoint, RfC),
    by = c("chemical" = "Chemical")
  ) %>%
  mutate(
    HQ = if_else(!is.na(RfC), concentration_total / RfC, NA_real_)
  )

# --- 4. Calculate ozone HQs separately for the 3 scenarios ---
ozone_hq <- ozone_final %>%
  mutate(Chemical = "ozone") %>%               # add Chemical column for joining
  left_join(
    rfcs_acute %>% filter(Chemical == "ozone") %>% select(Chemical, Endpoint, RfC),
    by = "Chemical"
  ) %>%
  mutate(
    HQ_time = if_else(!is.na(RfC), ozone_time_matched / RfC, NA_real_),
    HQ_p50  = if_else(!is.na(RfC), ozone_p50 / RfC, NA_real_),
    HQ_p95  = if_else(!is.na(RfC), ozone_p95 / RfC, NA_real_)
  ) %>%
  select(
    Date, Hour, Endpoint, RfC,
    ozone_time_matched, ozone_p50, ozone_p95,
    HQ_time, HQ_p50, HQ_p95
  )
```



#calculate HIs
```{r}
# Sum HQs for chemicals by timestamp, site, distance, and endpoint
chemical_hi <- chemical_hq %>%
  group_by(`Start Date Time`, `Source Wellpad`, `Sampling Site`, distance_ft, Operation, Endpoint) %>%
  summarise(
    HI_chemicals = sum(HQ, na.rm = TRUE),
    .groups = "drop"
  )

# Add Date and Hour to chemical HI for matching to ozone
chemical_hi <- chemical_hi %>%
  mutate(
    Date = as.Date(`Start Date Time`),
    Hour = hour(`Start Date Time`)
  )

# Join ozone HQs to chemical HI
ozone_hq <- ozone_hq %>%
  mutate(Hour = as.numeric(Hour))

hi_combined <- chemical_hi %>%
  left_join(
    ozone_hq %>%
      rename(
        HQ_ozone_time  = HQ_time,
        HQ_ozone_p50   = HQ_p50,
        HQ_ozone_p95   = HQ_p95
      ),
    by = c("Date", "Hour", "Endpoint")
  )

#calculate total HI for each ozone scenario
hi_combined <- hi_combined %>%
  mutate(
    HQ_ozone_time = dplyr::coalesce(HQ_ozone_time, 0),
    HQ_ozone_p50  = dplyr::coalesce(HQ_ozone_p50, 0),
    HQ_ozone_p95  = dplyr::coalesce(HQ_ozone_p95, 0)
  )

hi_combined <- hi_combined %>%
  mutate(
    HI_total_time = HI_chemicals + HQ_ozone_time,
    HI_total_p50  = HI_chemicals + HQ_ozone_p50,
    HI_total_p95  = HI_chemicals + HQ_ozone_p95
  )

#clean up
hi_final <- hi_combined %>%
  filter(!is.na(Endpoint)) %>%   # remove rows with missing endpoints
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    HI_chemicals,
    HI_total_time,
    HI_total_p50,
    HI_total_p95
  )

write.csv(hi_final, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/acute_HI.csv")
```



#plot HI by distance for an example trigger event
```{r}
#example for one trigger event
plot_df <- hi_final %>%
  filter(
    `Source Wellpad` == "NWB",
    `Sampling Site` == "UT2",
    format(`Start Date Time`, "%Y-%m-%d %H:%M") == "2021-09-18 08:22",
    Endpoint != "immunological"
  )

hi_distance_ex = ggplot(plot_df, aes(x = distance_ft, y = HI_total_time, color = Endpoint)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Distance (ft)",
    y = "HI (Time Matched Ozone)",
    color = "Endpoint",
    title = "Hazard Index vs Distance"
  ) +
  theme_minimal() +
  geom_hline(yintercept = 1, linetype = "dashed")

hi_distance_ex

ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/acute/hi_distance_example.png",
  plot = hi_distance_ex,
  width = 8,
  height = 6,
  dpi = 300
)
```



#plot HI vs. distance for each endpoint, colored/multiple lines with different ozone scenarios averaged (and faceted) across O&G operation
```{r}
#reshape to long
hi_long <- hi_final %>%
  pivot_longer(
    cols = c(HI_total_time, HI_total_p50, HI_total_p95),
    names_to = "ozone_scenario",
    values_to = "HI"
  )

#change ozone scenario names for graphing
hi_long <- hi_long %>%
  mutate(
    ozone_scenario = recode(
      ozone_scenario,
      "HI_total_time" = "Time-matched",
      "HI_total_p50"  = "Ozone P50",
      "HI_total_p95"  = "Ozone P95"
    )
  )

#summarize across Operations
hi_summary <- hi_long %>%
  group_by(Operation, Endpoint, distance_ft, ozone_scenario) %>%
  summarise(
    HI_mean   = mean(HI, na.rm = TRUE),
    HI_median = median(HI, na.rm = TRUE),
    HI_min    = min(HI, na.rm = TRUE),
    HI_max    = max(HI, na.rm = TRUE),
    HI_sd     = sd(HI, na.rm = TRUE),
    n         = sum(!is.na(HI)),
    HI_se     = HI_sd / sqrt(n),
    .groups = "drop"
  )


#plot of each Endpoint colored by ozone senario and faceted by Operation
plots_by_endpoint <- lapply(unique(hi_summary$Endpoint), function(ep) {
  
  df <- hi_summary %>%
    filter(Endpoint == ep)
  
  x_breaks <- sort(unique(df$distance_ft))
  x_breaks_thinned <- x_breaks[seq(1, length(x_breaks), by = 2)]
  
  ggplot(
    df,
    aes(
      x = distance_ft,
      y = HI_mean,
      color = ozone_scenario,
      group = ozone_scenario
    )
  ) +
    geom_line(size = 1) +
    geom_point() +
    geom_errorbar(
      aes(
        ymin = HI_mean - HI_se,
        ymax = HI_mean + HI_se
      ),
      width = 50,
      alpha = 0.6
    ) +
    ggh4x::facet_wrap2(~ Operation, axes = "all") +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_x_continuous(breaks = x_breaks_thinned) +   
    labs(
      title = paste("Hazard Index vs Distance -", ep),
      x = "Distance (ft)",
      y = "Mean Hazard Index",
      color = "Ozone Scenario"
    ) +
    theme_minimal()
})

#view plots
plots_by_endpoint[[1]]
plots_by_endpoint[[2]]
plots_by_endpoint[[3]]
plots_by_endpoint[[4]]
plots_by_endpoint[[5]]
plots_by_endpoint[[6]]
plots_by_endpoint[[7]]
plots_by_endpoint[[8]]
plots_by_endpoint[[9]]
plots_by_endpoint[[10]]
plots_by_endpoint[[11]]

#save plots
# get endpoint names
endpoint_names <- unique(hi_summary$Endpoint)

# loop and save
for (i in seq_along(plots_by_endpoint)) {
  
  # clean endpoint name for filename
  ep_name <- endpoint_names[i] %>%
    stringr::str_replace_all("[^A-Za-z0-9]", "_")
  
  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/acute",
      paste0("HI_dist_operation_", ep_name, ".png")
    ),
    plot = plots_by_endpoint[[i]],
    width = 10,
    height = 6,
    dpi = 300
  )
}

#save summary info for reporting
write.csv(hi_summary, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/HI_summary.csv")
```



## what chemicals are contributing across distance (where ozone contributes, looking at P95 scenario) - stacked bar graphs


#combining HQ dataset for ozone and chemicals (VOCs)
```{r}
# --- chemical HQs ---
chem_df <- chemical_hq %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    chemical,
    HQ
  )

# --- ozone HQ (P95) expanded regionally ---
ozone_hq_p95 <- ozone_hq %>%
  select(Date, Hour, Endpoint, HQ_p95) %>%
  rename(HQ = HQ_p95) %>%
  mutate(chemical = "ozone")

structure_df <- chemical_hq %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint
  ) %>%
  distinct() %>%
  mutate(
    Date = as.Date(`Start Date Time`),
    Hour = lubridate::hour(`Start Date Time`)
  )

ozone_expanded <- structure_df %>%
  left_join(ozone_hq_p95, by = c("Date", "Hour", "Endpoint")) %>%
  filter(!is.na(HQ)) %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    chemical,
    HQ
  )

# --- combine everything ---
hq_all <- bind_rows(chem_df, ozone_expanded)
```


#calculating average (mean) HQ per distance for each endpoint, operation, and distance
```{r}
hq_all_clean <- hq_all %>%
  filter(!is.na(HQ))

hq_plot_df <- hq_all_clean %>%
  group_by(Operation, Endpoint, distance_ft, chemical) %>%
  summarise(
    
    n = sum(!is.na(HQ)),
    
    HQ_mean = mean(HQ, na.rm = TRUE),
    HQ_median = median(HQ, na.rm = TRUE),
    
    HQ_sd = sd(HQ, na.rm = TRUE),
    
    HQ_se = ifelse(n > 0, HQ_sd / sqrt(n), NA_real_),
    
    HQ_min = min(HQ, na.rm = TRUE),
    HQ_max = max(HQ, na.rm = TRUE),
    
    .groups = "drop"
  )

#QC - stacked HQ equals HI (ok if these are 0/near zero)
hi_check <- hq_plot_df %>%
  group_by(Operation, Endpoint, distance_ft) %>%
  summarise(
    HI_from_stack = sum(HQ_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    hi_final %>%
      group_by(Operation, Endpoint, distance_ft) %>%
      summarise(
        HI_actual = mean(HI_total_p95, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("Operation", "Endpoint", "distance_ft")
  ) %>%
  mutate(diff = HI_from_stack - HI_actual)

summary(hi_check$diff)
```


#stacked bar plots
```{r}
plots_by_endpoint <- lapply(unique(hq_plot_df$Endpoint), function(ep) {
  
  df <- hq_plot_df %>%
    filter(Endpoint == ep)
  
  ggplot(df, aes(
    x = factor(distance_ft),
    y = HQ_mean,
    fill = chemical
  )) +
    
    geom_col(position = "stack", width = 0.8) +
    
    facet_wrap(~ Operation) +
    
    geom_hline(yintercept = 1, linetype = "dashed") +
    
    labs(
      title = paste("Stacked HQ Contributions by Distance -", ep),
      x = "Distance (ft)",
      y = "Mean Hazard Quotient (stacked = total HI)",
      fill = "Chemical"
    ) +
    
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
})

#view plots
plots_by_endpoint[[1]]
plots_by_endpoint[[2]]
plots_by_endpoint[[3]]
plots_by_endpoint[[4]]
plots_by_endpoint[[5]]
plots_by_endpoint[[6]]
plots_by_endpoint[[7]]
plots_by_endpoint[[8]]
plots_by_endpoint[[9]]
plots_by_endpoint[[10]]
plots_by_endpoint[[11]]

#save plots
endpoint_names <- unique(hq_plot_df$Endpoint)

for (i in seq_along(plots_by_endpoint)) {
  
  ep_name <- endpoint_names[i] %>%
    stringr::str_replace_all("[^A-Za-z0-9]", "_")
  
  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/acute",
      paste0("HQ_dist_operation_bar_", ep_name, ".png")
    ),
    plot = plots_by_endpoint[[i]],
    width = 11,
    height = 7,
    dpi = 300
  )
}

#save HQ summary data
write.csv(hq_plot_df, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/HQ_summary.csv")
```








