---
title: "Acute aggregate HI analysis with background"
output: html_document
date: "2026-04-24"
---


#load libraries 
```{r}
library("dplyr")
library("tidyr")
library("Rmisc")
library("terra")
library("readxl")
library("tidyverse")
library("stringr")
library("lubridate")
library("ggh4x")
library("patchwork")
library("cowplot")
library("grid")
detach(package:Rmisc)
detach(package:plyr)
```



#load chronic "background" data
```{r}
file1 <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/chronic_spatial_trend_monthly_gapfilled_with_bkg.xlsx"

sheets <- excel_sheets(file1)

chronic_list <- lapply(sheets, function(s) {

  df <- read_excel(file1, sheet = s)

  df_long <- df %>%
    pivot_longer(
      cols = starts_with("Distance_"),
      names_to = "distance_ft",
      values_to = "concentration_ppb"
    ) %>%
    mutate(
      distance_ft = as.numeric(str_extract(distance_ft, "\\d+")),
      chemical = s
    )

  return(df_long)
})

chronic <- bind_rows(chronic_list)

chronic <- chronic %>%
  mutate(chemical = recode(
    chemical,
    "aromatic medium" = "aromatic_medium",
    "aliphatic low"   = "aliphatic_low",
    "nonane" = "n_nonane",
    "decane" = "n_decane"
  ))

## save dataframe for reporting ##

write.csv(chronic, file = "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/acute_chronicmonthly_clean.csv")
```



#load acute trigger cannister data
```{r}
file2 <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/Broomfield-Triggered Canisters-VOC concentration vs Distances.xlsx"

acute_raw <- read_excel(file2)

meta_cols <- c(
  "Start Date Time",
  "Stop Date Time",
  "Source Wellpad",
  "Operation",
  "Sampling Site",
  "Distance (site, pad) ft",
  "Note"
)

acute <- acute_raw %>%
  select(
    all_of(meta_cols),
    matches("\\d+ft ppb$")   # keeps things like "benzene 500ft ppb"
  )

# reshape to long format
acute_long <- acute %>%
  pivot_longer(
    cols = -all_of(meta_cols),
    names_to = "chem_dist",
    values_to = "concentration_ppb"
  )

#separate chemical and distance
acute_long <- acute_long %>%
  mutate(
    chemical = str_extract(chem_dist, "^[^ ]+"),
    distance_ft = as.numeric(str_extract(chem_dist, "\\d+(?=ft)"))
  ) %>%
  select(-chem_dist)
```



#filter dataset and combine chemicals into groups 
```{r}
group_vars <- c(
  "Start Date Time",
  "Stop Date Time",
  "Source Wellpad",
  "Operation",
  "Sampling Site",
  "Distance (site, pad) ft",
  "distance_ft"
)

acute_long <- acute_long %>%
  rename(concentration = concentration_ppb)

#group chemical categories
## xylenes
xylenes_df <- acute_long %>%
  filter(chemical %in% c("o_xylene", "m_p_xylene")) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(concentration = sum(concentration, na.rm = TRUE)) %>%
  mutate(chemical = "xylenes")

## aliphatic_low
aliphatic_low_chems <- c(
  "i_pentane", "n_pentane", "n_hexane", "n_heptane", "n_octane",
  "cyclopentane", "cyclohexane", "methylcyclohexane",
  "2_2_4_trimethylpentane", "2_3_4_trimethylpentane",
  "2_4_dimethylpentane", "2_3_dimethylpentane",
  "2_methylhexane", "3_methylhexane",
  "2_methylheptane", "3_methylheptane",
  "t_2_pentene", "1_pentene", "c_2_pentene",
  "isoprene"
)

aliphatic_low_df <- acute_long %>%
  filter(chemical %in% aliphatic_low_chems) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(concentration = sum(concentration, na.rm = TRUE)) %>%
  mutate(chemical = "aliphatic_low")

## aromatic_medium
aromatic_medium_chems <- c(
  "i_propylbenzene", "n_propylbenzene",
  "2_ethyltoluene", "3_ethyltoluene", "4_ethyltoluene",
  "1_2_3_trimethylbenzene", "1_2_4_trimethylbenzene", "1_3_5_trimethylbenzene",
  "1_3_diethylbenzene", "1_4_diethylbenzene"
)

aromatic_medium_df <- acute_long %>%
  filter(chemical %in% aromatic_medium_chems) %>%
  group_by(across(all_of(group_vars))) %>%
  summarise(concentration = sum(concentration, na.rm = TRUE)) %>%
  mutate(chemical = "aromatic_medium")

#remove unwanted chems
drop_chems <- c(
  # components now grouped
  "o_xylene", "m_p_xylene",
  aliphatic_low_chems,
  aromatic_medium_chems,

  # explicitly remove
  "methane", "ethane", "propane",
  "i_butane", "n_butane",
  "ethene", "ethyne",
  "t_2_butene", "1_butene", "c_2_butene",
  "NMVOC"
)

acute_filtered <- acute_long %>%
  filter(!chemical %in% drop_chems)

#final clean dataset
acute_clean <- bind_rows(
  acute_filtered,
  xylenes_df,
  aliphatic_low_df,
  aromatic_medium_df
)

#checks
# check new chemical list
sort(unique(acute_clean$chemical))

# make sure grouped ones exist
table(acute_clean$chemical %in% c("xylenes", "aliphatic_low", "aromatic_medium"))

# check no duplicates per group
dup_check <- acute_clean %>%
  group_by(across(all_of(group_vars)), chemical) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1)

nrow(dup_check)

setdiff(group_vars, names(acute_clean))

setdiff(aliphatic_low_chems, unique(acute_long$chemical))

## save dataframe for reporting ##

write.csv(acute_clean, file = "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/acute_trigger_clean.csv")
```



#calculate actual, 50th, and 95th percentile ozone concentrations
```{r}
ozone = read.csv("~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/ozone_hourly_2018-2023.csv")

#prepare data with time stamps
ozone_clean <- ozone %>%
  mutate(
    Date = as.Date(Date.Local),
    Hour = format(as.POSIXct(Time.Local, format = "%H:%M"), "%H"),
    ozone_ppb = Sample.Measurement * 1000   # ppm → ppb
  ) %>%
  filter(lubridate::year(Date) >= 2019 & lubridate::year(Date) <= 2022)

acute_time_index <- acute_clean %>%
  mutate(
    Date = as.Date(`Start Date Time`),
    Hour = format(as.POSIXct(`Start Date Time`), "%H")
  ) %>%
  distinct(Date, Hour)

#time-matched ozone concentrations (scenario 1)
ozone_time_matched <- ozone_clean %>%
  semi_join(acute_time_index, by = c("Date", "Hour")) %>%
  group_by(Date, Hour) %>%
  summarise(
    ozone_time_matched = median(ozone_ppb, na.rm = TRUE),
    .groups = "drop"
  )

#50th and 95th percentile concentrations
ozone_p50 <- median(ozone_clean$ozone_ppb, na.rm = TRUE)
ozone_p95 <- quantile(ozone_clean$ozone_ppb, 0.95, na.rm = TRUE)

#final ozone in one df
ozone_final <- ozone_time_matched %>%
  mutate(
    ozone_p50 = ozone_p50,
    ozone_p95 = ozone_p95
  )

#checks
# should match acute timestamps
nrow(ozone_final)
nrow(acute_time_index)

# check missing ozone values
sum(is.na(ozone_final$ozone_time_matched))

## save ozone data for reporting ##
write.csv(ozone_final, file = "~/CO_CRA/CO_CRA_aggregate/data/broomfield/acute/acute_ozone_clean.csv")

```


#match chemicl names to rfcs 
```{r}
rfcs = read.csv("~/CO_CRA/CO_CRA_aggregate/data/RfCs_5-2026.csv")

rfcs_acute <- rfcs %>%
  dplyr::filter(Exposure == "acute")

rfcs_acute$Chemical[rfcs_acute$Chemical == "Benzene"] <- "benzene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Toluene"] <- "toluene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Ethylbenzene"] <- "ethylbenzene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Xylenes"] <- "xylenes"
rfcs_acute$Chemical[rfcs_acute$Chemical == "n-nonane"] <- "n_nonane"
rfcs_acute$Chemical[rfcs_acute$Chemical == "n-decane"] <- "n_decane"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Aliphatic low"] <- "aliphatic_low"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Aromatic medium"] <- "aromatic_medium"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Styrene"] <- "styrene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Trichloroethylene (TCE)"] <- "c2hcl3"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Tetrachloroethylene (PCE)"] <- "c2cl4"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Propene"] <- "propene"
rfcs_acute$Chemical[rfcs_acute$Chemical == "Ozone"] <- "ozone"

#check
setdiff(acute_clean$chemical, rfcs_acute$Chemical)
setdiff(rfcs_acute$Chemical, acute_clean$chemical)
```



#calculate HQs
```{r}
# --- 1. Prepare chronic dataset to match acute ---
chronic_ready <- chronic %>%
  rename(concentration = concentration_ppb) %>%
  mutate(
    Operation = NA_character_,
    distance_ft = as.numeric(distance_ft)
  ) %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    chemical,
    concentration,
    Operation
  )

# --- 2. Combine acute + chronic concentrations ---
combined_df <- bind_rows(
  acute_clean %>%
    select(
      `Start Date Time`,
      `Source Wellpad`,
      `Sampling Site`,
      distance_ft,
      chemical,
      concentration,
      Operation
    ),
  chronic_ready
)

# Sum concentrations per timestamp, site, chemical, distance
combined_df <- combined_df %>%
  group_by(`Start Date Time`, `Source Wellpad`, `Sampling Site`, distance_ft, chemical) %>%
  summarise(
    concentration_total = sum(concentration, na.rm = TRUE),
    Operation = dplyr::first(na.omit(Operation)), # take acute Operation if available
    .groups = "drop"
  )

# --- 3. Calculate chemical HQs ---
chemical_hq <- combined_df %>%
  filter(chemical != "ozone") %>%
  left_join(
    rfcs_acute %>% select(Chemical, Endpoint, RfC),
    by = c("chemical" = "Chemical")
  ) %>%
  mutate(
    HQ = if_else(!is.na(RfC), concentration_total / RfC, NA_real_)
  )

# --- 4. Calculate ozone HQs separately for the 3 scenarios ---
ozone_hq <- ozone_final %>%
  mutate(Chemical = "ozone") %>%               # add Chemical column for joining
  left_join(
    rfcs_acute %>% filter(Chemical == "ozone") %>% select(Chemical, Endpoint, RfC),
    by = "Chemical"
  ) %>%
  mutate(
    HQ_time = if_else(!is.na(RfC), ozone_time_matched / RfC, NA_real_),
    HQ_p50  = if_else(!is.na(RfC), ozone_p50 / RfC, NA_real_),
    HQ_p95  = if_else(!is.na(RfC), ozone_p95 / RfC, NA_real_)
  ) %>%
  select(
    Date, Hour, Endpoint, RfC,
    ozone_time_matched, ozone_p50, ozone_p95,
    HQ_time, HQ_p50, HQ_p95
  )

## save data for reporting, with XX ##

write.csv(chemical_hq, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/acute_HQs_chemicals.csv")
write.csv(ozone_hq, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/acute_HQs_ozone.csv")

```



#calculate HIs
```{r}
# Sum HQs for chemicals by timestamp, site, distance, and endpoint
chemical_hi <- chemical_hq %>%
  group_by(`Start Date Time`, `Source Wellpad`, `Sampling Site`, distance_ft, Operation, Endpoint) %>%
  summarise(
    HI_chemicals = sum(HQ, na.rm = TRUE),
    .groups = "drop"
  )

# Add Date and Hour to chemical HI for matching to ozone
chemical_hi <- chemical_hi %>%
  mutate(
    Date = as.Date(`Start Date Time`),
    Hour = hour(`Start Date Time`)
  )

# Join ozone HQs to chemical HI
ozone_hq <- ozone_hq %>%
  mutate(Hour = as.numeric(Hour))

hi_combined <- chemical_hi %>%
  left_join(
    ozone_hq %>%
      rename(
        HQ_ozone_time  = HQ_time,
        HQ_ozone_p50   = HQ_p50,
        HQ_ozone_p95   = HQ_p95
      ),
    by = c("Date", "Hour", "Endpoint")
  )

#calculate total HI for each ozone scenario
hi_combined <- hi_combined %>%
  mutate(
    HQ_ozone_time = dplyr::coalesce(HQ_ozone_time, 0),
    HQ_ozone_p50  = dplyr::coalesce(HQ_ozone_p50, 0),
    HQ_ozone_p95  = dplyr::coalesce(HQ_ozone_p95, 0)
  )

hi_combined <- hi_combined %>%
  mutate(
    HI_total_time = HI_chemicals + HQ_ozone_time,
    HI_total_p50  = HI_chemicals + HQ_ozone_p50,
    HI_total_p95  = HI_chemicals + HQ_ozone_p95
  )

#clean up
hi_final <- hi_combined %>%
  filter(!is.na(Endpoint)) %>%   # remove rows with missing endpoints
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    HI_chemicals,
    HI_total_time,
    HI_total_p50,
    HI_total_p95
  )

write.csv(hi_final, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/acute_HI_wbkg.csv")
```




#plot HI vs. distance for each endpoint, colored/multiple lines with different ozone scenarios averaged (and faceted) across O&G operation (not using in MS)
```{r}
#reshape to long
hi_long <- hi_final %>%
  pivot_longer(
    cols = c(HI_total_time, HI_total_p50, HI_total_p95),
    names_to = "ozone_scenario",
    values_to = "HI"
  )

#change ozone scenario names for graphing
hi_long <- hi_long %>%
  mutate(
    ozone_scenario = recode(
      ozone_scenario,
      "HI_total_time" = "Time-matched",
      "HI_total_p50"  = "Ozone P50",
      "HI_total_p95"  = "Ozone P95"
    )
  )

#summarize across Operations
hi_summary <- hi_long %>%
  group_by(Operation, Endpoint, distance_ft, ozone_scenario) %>%
  summarise(
    HI_mean   = mean(HI, na.rm = TRUE),
    HI_median = median(HI, na.rm = TRUE),
    HI_min    = min(HI, na.rm = TRUE),
    HI_max    = max(HI, na.rm = TRUE),
    HI_sd     = sd(HI, na.rm = TRUE),
    n         = sum(!is.na(HI)),
    HI_se     = HI_sd / sqrt(n),
    .groups = "drop"
  )


#plot of each Endpoint colored by ozone senario and faceted by Operation
plots_by_endpoint <- lapply(unique(hi_summary$Endpoint), function(ep) {
  
  df <- hi_summary %>%
    filter(Endpoint == ep)
  
  x_breaks <- sort(unique(df$distance_ft))
  x_breaks_thinned <- x_breaks[seq(1, length(x_breaks), by = 2)]
  
  ggplot(
    df,
    aes(
      x = distance_ft,
      y = HI_mean,
      color = ozone_scenario,
      group = ozone_scenario
    )
  ) +
    geom_line(size = 1) +
    geom_point() +
    geom_errorbar(
      aes(
        ymin = HI_mean - HI_se,
        ymax = HI_mean + HI_se
      ),
      width = 50,
      alpha = 0.6
    ) +
    ggh4x::facet_wrap2(~ Operation, axes = "all") +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_x_continuous(breaks = x_breaks_thinned) +   
    labs(
      title = paste("Hazard Index vs Distance -", ep),
      x = "Distance from Wellpad (ft)",
      y = "Average Hazard Index",
      color = "Ozone Scenario"
    ) +
scale_color_manual(
  values = c(
    "Time-matched" = "#7F7F7F",
    "Ozone P50"    = "#6BAED6",
    "Ozone P95"    = "#08519C"
  )
) +
    theme_minimal()
})

#view plots
plots_by_endpoint[[1]]
plots_by_endpoint[[2]]
plots_by_endpoint[[3]]
plots_by_endpoint[[4]]
plots_by_endpoint[[5]]
plots_by_endpoint[[6]]
plots_by_endpoint[[7]]
plots_by_endpoint[[8]]
plots_by_endpoint[[9]]
plots_by_endpoint[[10]]
plots_by_endpoint[[11]]

#save plots
# get endpoint names
endpoint_names <- unique(hi_summary$Endpoint)


# loop and save
for (i in seq_along(plots_by_endpoint)) {
  
  # clean endpoint name for filename
  ep_name <- endpoint_names[i] %>%
    stringr::str_replace_all("[^A-Za-z0-9]", "_")
  
  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/acute",
      paste0("HI_dist_operation_wbkg_", ep_name, ".png")
    ),
    plot = plots_by_endpoint[[i]],
    width = 10,
    height = 6,
    dpi = 300
  )
}

#save summary info for reporting
write.csv(hi_summary, file = "~/CO_CRA/CO_CRA_aggregate/output/acute/HI_summary_wbkg.csv")

hi_plots_by_endpoint <- plots_by_endpoint


```




## what chemicals are contributing across distance (where ozone contributes, looking at P95 scenario) - stacked bar graphs


#combining HQ dataset for ozone and chemicals (VOCs)
```{r}
# --- chemical HQs ---
chem_df <- chemical_hq %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    chemical,
    HQ
  )

# --- ozone HQ (P95) expanded regionally ---
ozone_hq_p95 <- ozone_hq %>%
  select(Date, Hour, Endpoint, HQ_p95) %>%
  rename(HQ = HQ_p95) %>%
  mutate(chemical = "ozone")

structure_df <- chemical_hq %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint
  ) %>%
  distinct() %>%
  mutate(
    Date = as.Date(`Start Date Time`),
    Hour = lubridate::hour(`Start Date Time`)
  )

ozone_expanded <- structure_df %>%
  left_join(ozone_hq_p95, by = c("Date", "Hour", "Endpoint")) %>%
  filter(!is.na(HQ)) %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    chemical,
    HQ
  )

# --- combine everything ---
hq_all <- bind_rows(chem_df, ozone_expanded)
```



#calculating average (mean) HQ per distance for each endpoint, operation, and distance
```{r}
hq_all_clean <- hq_all %>%
  filter(!is.na(HQ))

hq_plot_df <- hq_all_clean %>%
  group_by(Operation, Endpoint, distance_ft, chemical) %>%
  summarise(
    
    n = sum(!is.na(HQ)),
    
    HQ_mean = mean(HQ, na.rm = TRUE),
    HQ_median = median(HQ, na.rm = TRUE),
    
    HQ_sd = sd(HQ, na.rm = TRUE),
    
    HQ_se = ifelse(n > 0, HQ_sd / sqrt(n), NA_real_),
    
    HQ_min = min(HQ, na.rm = TRUE),
    HQ_max = max(HQ, na.rm = TRUE),
    
    .groups = "drop"
  )

#QC - stacked HQ equals HI (ok if these are 0/near zero)
hi_check <- hq_plot_df %>%
  group_by(Operation, Endpoint, distance_ft) %>%
  summarise(
    HI_from_stack = sum(HQ_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    hi_final %>%
      group_by(Operation, Endpoint, distance_ft) %>%
      summarise(
        HI_actual = mean(HI_total_p95, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("Operation", "Endpoint", "distance_ft")
  ) %>%
  mutate(diff = HI_from_stack - HI_actual)

summary(hi_check$diff)
```



#stacked bar plots (ozone is for the p95 scenario)
```{r}
# -----------------------------
# Change chemical names
# -----------------------------

sort(unique(hq_plot_df$chemical))

hq_plot_df <- hq_plot_df %>%
  mutate(
    chemical = recode(
      chemical,
      "aliphatic_low"   = "Aliphatic low",
      "aromatic_medium" = "Aromatic medium",
      "benzene"         = "Benzene",
      "c2cl4"           = "PCE",
      "c2hcl3"          = "TCE",
      "ethylbenzene"    = "Ethylbenzene",
      "n_nonane"        = "n_Nonane",
      "Ozone"           = "Ozone (p95)",
      "ozone"           = "Ozone (p95)",
      "styrene"         = "Styrene",
      "toluene"         = "Toluene",
      "xylenes"         = "Xylenes"
    )
  )


# -----------------------------
# Define chemical colors
# -----------------------------

chemical_colors <- c(
  "Aliphatic low"   = "#FFFFD9",
  "Aromatic medium" = "#EDF8B1",
  "Benzene"         = "#D9E57A",
  "Ethylbenzene"    = "#C7E9B4",
  "n_Nonane"        = "#7FCDBB",
  "Ozone (p95)"     = "#6AB4A7",
  "PCE"             = "#41B6C4",
  "Styrene"         = "#1D91C0",
  "TCE"             = "#225EA8",
  "Toluene"         = "#253494",
  "Xylenes"         = "#081D58"
)


# -----------------------------
# Define legend labels
# -----------------------------

chemical_legend_labels <- c(
  "Aliphatic low"   = "Aliphatic low (Al)",
  "Aromatic medium" = "Aromatic medium (Am)",
  "Benzene"         = "Benzene (B)",
  "Ethylbenzene"    = "Ethylbenzene (E)",
  "n_Nonane"        = "n_Nonane (N)",
  "Ozone (p95)"     = "p95 Ozone (O)",
  "PCE"             = "PCE (P)",
  "Styrene"         = "Styrene (S)",
  "TCE"             = "TCE (T)",
  "Toluene"         = "Toluene (To)",
  "Xylenes"         = "Xylenes (X)"
)


# -----------------------------
# Set consistent chemical order
# -----------------------------

chemical_order <- names(chemical_colors)

hq_plot_df <- hq_plot_df %>%
  mutate(
    chemical = factor(
      chemical,
      levels = chemical_order
    )
  )


# -----------------------------
# Create one paneled plot per endpoint
# -----------------------------

endpoint_names <- unique(hq_plot_df$Endpoint)

plots_by_endpoint <- lapply(endpoint_names, function(ep) {

  endpoint_df <- hq_plot_df %>%
    filter(Endpoint == ep) %>%
    droplevels()

  operation_names <- unique(endpoint_df$Operation)

  # Number of operation panels and graph rows
  n_operations <- length(operation_names)
  n_rows <- ceiling(n_operations / 2)

  # Number of chemicals shown in this endpoint's legend
  n_legend_entries <- endpoint_df %>%
    filter(!is.na(chemical)) %>%
    distinct(chemical) %>%
    nrow()

  # Give the bottom row extra height when the legend is long
  bottom_row_height <- if (n_legend_entries >= 9) {
    1.55
  } else if (n_legend_entries >= 7) {
    1.30
  } else {
    1.10
  }

  # Relative heights for the graph rows
  row_heights <- c(
    rep(1, n_rows - 1),
    bottom_row_height
  )


  # -----------------------------
  # Calculate complete stacked HQ totals
  # -----------------------------

  endpoint_stack_totals <- endpoint_df %>%
    group_by(
      Operation,
      distance_ft
    ) %>%
    summarise(
      stacked_HQ = sum(
        HQ_mean,
        na.rm = TRUE
      ),
      .groups = "drop"
    )

  endpoint_y_max <- max(
    endpoint_stack_totals$stacked_HQ,
    na.rm = TRUE
  )

  # Add space above the tallest complete stack
  endpoint_y_limit <- endpoint_y_max * 1.05

  # Fallback if all values are missing or zero
  if (
    !is.finite(endpoint_y_limit) ||
      endpoint_y_limit <= 0
  ) {
    endpoint_y_limit <- 1
  }


  # -----------------------------
  # Create one plot per operation
  # -----------------------------

  operation_plots <- lapply(
    seq_along(operation_names),
    function(op_index) {

      op <- operation_names[op_index]

      operation_df <- endpoint_df %>%
        filter(Operation == op) %>%
        droplevels()

      ggplot(
        operation_df,
        aes(
          x = factor(distance_ft),
          y = HQ_mean,
          fill = chemical
        )
      ) +

        geom_col(
          position = "stack",
          width = 0.8
        ) +

        geom_hline(
          yintercept = 1,
          linetype = "dashed",
          linewidth = 0.6
        ) +

        scale_fill_manual(
          name = "Chemical",
          values = chemical_colors,
          breaks = chemical_order,
          labels = chemical_legend_labels,
          drop = TRUE
        ) +

        # Do not place limits directly in the scale because
        # that can remove portions of stacked bars
        scale_y_continuous(
          expand = expansion(
            mult = c(0, 0.02)
          )
        ) +

        # Apply displayed range after bars are stacked
        coord_cartesian(
          ylim = c(
            0,
            endpoint_y_limit
          )
        ) +

labs(
  # Keep A, B, C, etc.
  tag = LETTERS[op_index],

  # Remove operation title
  title = NULL,

  # Shared axis titles are added to the full figure
  x = NULL,
  y = NULL
) +

        guides(
          fill = guide_legend(
            ncol = 1,
            byrow = TRUE
          )
        ) +

        theme_minimal(
          base_size = 11
        ) +

        theme(
          plot.title = element_text(
            hjust = 0.5,
            face = "bold",
            size = 12,
            lineheight = 0.95,
            margin = margin(
              b = 6
            )
          ),

          # Panel label appearance and position
          plot.tag = element_text(
            face = "bold",
            size = 11,
            hjust = 0,
            vjust = 1
          ),

          plot.tag.position = c(
            -0.06,
            1
          ),

          axis.text.x = element_text(
            angle = 45,
            hjust = 1,
            vjust = 1,
            size = 9.5
          ),

          axis.text.y = element_text(
            size = 10
          ),

          # Remove individual panel axis titles
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),

          legend.title = element_text(
            face = "bold",
            size = 11
          ),

          legend.text = element_text(
            size = 10
          ),

          legend.key.size = grid::unit(
            0.44,
            "cm"
          ),

          legend.key.height = grid::unit(
            0.44,
            "cm"
          ),

          legend.key.width = grid::unit(
            0.44,
            "cm"
          ),

          legend.spacing.y = grid::unit(
            0.03,
            "cm"
          ),

          legend.margin = margin(
            t = 0,
            r = 0,
            b = 0,
            l = 0
          ),

          panel.grid.minor = element_blank(),

          # Slightly more space for panel tag
          plot.margin = margin(
            t = 8,
            r = 5,
            b = 5,
            l = 8
          )
        )
    }
  )

  names(operation_plots) <- operation_names


  # -----------------------------
  # Add legend placeholder
  # -----------------------------

  plot_list_with_legend <- c(
    operation_plots,
    list(
      patchwork::guide_area()
    )
  )


  # -----------------------------
  # Assemble two-column graph section
  # -----------------------------

graph_section <- patchwork::wrap_plots(
  plotlist = plot_list_with_legend,
  ncol = 2,
  guides = "collect",
  heights = row_heights
) &

  theme(
    legend.position = "right",
    legend.justification = "center",
    legend.box = "vertical"
  )


  # -----------------------------
  # Create one shared x-axis title
  # -----------------------------

  shared_x_axis_title <- patchwork::wrap_elements(
    full = grid::textGrob(
      label = "Distance from Well Pad (ft)",
      x = 0.5,
      y = 0.65,
      just = "center",
      gp = grid::gpar(
        fontsize = 11.5
      )
    ),
    ignore_tag = TRUE
  )


  # -----------------------------
  # Create one shared y-axis title
  # -----------------------------

  shared_y_axis_title <- patchwork::wrap_elements(
    full = grid::textGrob(
      label = "Average Hazard Quotient (HQ)",
      x = 0.55,
      y = 0.5,
      just = "center",
      rot = 90,
      gp = grid::gpar(
        fontsize = 11.5
      )
    ),
    ignore_tag = TRUE
  )


  # -----------------------------
  # Add shared y-axis title beside graph section
  # -----------------------------

  graph_with_y_title <- (
    shared_y_axis_title |
      graph_section
  ) +

    patchwork::plot_layout(
      widths = c(
        0.045,
        1
      )
    )


  # -----------------------------
  # Add shared x-axis title beneath complete figure
  # -----------------------------

  endpoint_plot <- (
    graph_with_y_title /
      shared_x_axis_title
  ) +

    patchwork::plot_layout(
      heights = c(
        1,
        0.035
      )
    )

  endpoint_plot
})


# -----------------------------
# View plots
# -----------------------------

plots_by_endpoint[[1]]
plots_by_endpoint[[2]]
plots_by_endpoint[[3]]
plots_by_endpoint[[4]]
plots_by_endpoint[[5]]
plots_by_endpoint[[6]]
plots_by_endpoint[[7]]
plots_by_endpoint[[8]]
plots_by_endpoint[[9]]
plots_by_endpoint[[10]]
plots_by_endpoint[[11]]


# -----------------------------
# Save plots
# -----------------------------

for (i in seq_along(plots_by_endpoint)) {

  ep_name <- endpoint_names[i] %>%
    stringr::str_replace_all(
      "[^A-Za-z0-9]",
      "_"
    )

  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/acute",
      paste0(
        "HQ_dist_operation_bar_wbkg_",
        ep_name,
        ".png"
      )
    ),

    plot = plots_by_endpoint[[i]],

    width = 7,
    height = 10,
    units = "in",
    dpi = 400
  )
}


# -----------------------------
# Save HQ summary data
# -----------------------------

write.csv(
  hq_plot_df,
  file = paste0(
    "~/CO_CRA/CO_CRA_aggregate/output/acute/",
    "HQ_summary_wbkg.csv"
  ),
  row.names = FALSE
)


# -----------------------------
# Retain plot object under original name
# -----------------------------

hq_plots_by_endpoint <- plots_by_endpoint
```




#save HQ_summary_wbkg file that includes all ozone scenarios for supplemental in the MS
```{r}
# ------------------------------------------------------------
# 1. Prepare the chemical HQ data
#
# These are the non-ozone chemical HQ values already calculated
# in chemical_hq.
# ------------------------------------------------------------

chemical_hq_allozone_raw <- chemical_hq %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    chemical,
    HQ
  ) %>%
  filter(
    !is.na(Operation),
    !is.na(Endpoint),
    !is.na(distance_ft),
    !is.na(chemical),
    !is.na(HQ)
  )


# ------------------------------------------------------------
# 2. Summarize the chemical HQs
#
# Each chemical appears only once per operation, endpoint,
# and distance.
# ------------------------------------------------------------

chemical_hq_allozone_summary <- chemical_hq_allozone_raw %>%
  group_by(
    Operation,
    Endpoint,
    distance_ft,
    chemical
  ) %>%
  summarise(
    n = sum(
      !is.na(HQ)
    ),

    HQ_mean = mean(
      HQ,
      na.rm = TRUE
    ),

    HQ_median = median(
      HQ,
      na.rm = TRUE
    ),

    HQ_sd = sd(
      HQ,
      na.rm = TRUE
    ),

    HQ_se = if_else(
      n > 0,
      HQ_sd / sqrt(n),
      NA_real_
    ),

    HQ_min = min(
      HQ,
      na.rm = TRUE
    ),

    HQ_max = max(
      HQ,
      na.rm = TRUE
    ),

    .groups = "drop"
  )


# ------------------------------------------------------------
# 3. Create the event, operation, endpoint, and distance
#    structure onto which ozone HQs will be joined
# ------------------------------------------------------------

ozone_structure_allozone <- chemical_hq %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint
  ) %>%
  filter(
    !is.na(Operation),
    !is.na(Endpoint),
    !is.na(distance_ft)
  ) %>%
  distinct() %>%
  mutate(
    Date = as.Date(
      `Start Date Time`
    ),

    Hour = lubridate::hour(
      `Start Date Time`
    )
  )


# ------------------------------------------------------------
# 4. Reshape the ozone HQ data
#
# The scenario names are placed directly into the chemical
# column so they can be appended to the chemical HQ summary.
# ------------------------------------------------------------

ozone_hq_allozone_long <- ozone_hq %>%
  select(
    Date,
    Hour,
    Endpoint,
    HQ_time,
    HQ_p50,
    HQ_p95
  ) %>%
  mutate(
    Hour = as.numeric(
      Hour
    )
  ) %>%
  pivot_longer(
    cols = c(
      HQ_time,
      HQ_p50,
      HQ_p95
    ),

    names_to = "chemical",
    values_to = "HQ"
  ) %>%
  mutate(
    chemical = recode(
      chemical,
      "HQ_time" = "Ozone (time-matched)",
      "HQ_p50"  = "Ozone (p50)",
      "HQ_p95"  = "Ozone (p95)"
    )
  )


# ------------------------------------------------------------
# 5. Expand ozone HQs to each event, operation, endpoint,
#    and distance
# ------------------------------------------------------------

ozone_hq_allozone_expanded <- ozone_structure_allozone %>%
  left_join(
    ozone_hq_allozone_long,

    by = c(
      "Date",
      "Hour",
      "Endpoint"
    )
  ) %>%
  filter(
    !is.na(HQ),
    !is.na(chemical)
  ) %>%
  select(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    Operation,
    Endpoint,
    chemical,
    HQ
  )


# ------------------------------------------------------------
# 6. Summarize the three ozone HQ scenarios
#
# This produces one summary row for each ozone scenario by
# operation, endpoint, and distance.
# ------------------------------------------------------------

ozone_hq_allozone_summary <- ozone_hq_allozone_expanded %>%
  group_by(
    Operation,
    Endpoint,
    distance_ft,
    chemical
  ) %>%
  summarise(
    n = sum(
      !is.na(HQ)
    ),

    HQ_mean = mean(
      HQ,
      na.rm = TRUE
    ),

    HQ_median = median(
      HQ,
      na.rm = TRUE
    ),

    HQ_sd = sd(
      HQ,
      na.rm = TRUE
    ),

    HQ_se = if_else(
      n > 0,
      HQ_sd / sqrt(n),
      NA_real_
    ),

    HQ_min = min(
      HQ,
      na.rm = TRUE
    ),

    HQ_max = max(
      HQ,
      na.rm = TRUE
    ),

    .groups = "drop"
  )


# ------------------------------------------------------------
# 7. Combine chemical and ozone summaries
#
# The ozone scenarios are appended as three additional
# "chemical" categories.
# ------------------------------------------------------------

hq_summary_allozone <- bind_rows(
  chemical_hq_allozone_summary,
  ozone_hq_allozone_summary
)


# ------------------------------------------------------------
# 8. Clean the non-ozone chemical names
# ------------------------------------------------------------

hq_summary_allozone <- hq_summary_allozone %>%
  mutate(
    chemical = recode(
      chemical,
      "aliphatic_low"   = "Aliphatic low",
      "aromatic_medium" = "Aromatic medium",
      "benzene"         = "Benzene",
      "c2cl4"           = "PCE",
      "c2hcl3"          = "TCE",
      "ethylbenzene"    = "Ethylbenzene",
      "n_nonane"        = "n_Nonane",
      "n_decane"        = "n_Decane",
      "styrene"         = "Styrene",
      "toluene"         = "Toluene",
      "xylenes"         = "Xylenes"
    )
  ) %>%
  arrange(
    Operation,
    Endpoint,
    distance_ft,
    chemical
  )


# ------------------------------------------------------------
# 9. Check the final chemical categories
# ------------------------------------------------------------

sort(
  unique(
    hq_summary_allozone$chemical
  )
)


# The ozone categories should include:
#
# "Ozone (time-matched)"
# "Ozone (p50)"
# "Ozone (p95)"


# ------------------------------------------------------------
# 10. Check that all three ozone scenarios are present for
#     each operation, endpoint, and distance
# ------------------------------------------------------------

ozone_scenario_check <- hq_summary_allozone %>%
  filter(
    chemical %in% c(
      "Ozone (time-matched)",
      "Ozone (p50)",
      "Ozone (p95)"
    )
  ) %>%
  count(
    Operation,
    Endpoint,
    distance_ft,
    name = "n_ozone_scenarios"
  )


# Ideally, every value should equal 3
table(
  ozone_scenario_check$n_ozone_scenarios
)


# Display any combinations that do not have all three scenarios
ozone_scenario_check %>%
  filter(
    n_ozone_scenarios != 3
  )


# ------------------------------------------------------------
# 11. Confirm that the chemical rows were not repeated
#
# Each chemical/operation/endpoint/distance combination should
# appear exactly once.
# ------------------------------------------------------------

duplicate_summary_check <- hq_summary_allozone %>%
  count(
    Operation,
    Endpoint,
    distance_ft,
    chemical,
    name = "n_rows"
  ) %>%
  filter(
    n_rows > 1
  )


nrow(
  duplicate_summary_check
)

# Expected result: 0


# ------------------------------------------------------------
# 12. Optional check of the number of summary rows by chemical
# ------------------------------------------------------------

hq_summary_allozone %>%
  count(
    chemical,
    name = "n_summary_rows"
  ) %>%
  arrange(
    chemical
  )


# ------------------------------------------------------------
# 13. Quality-control check:
#     p95 chemical HQs + p95 ozone should reproduce the
#     mean p95 HI for each operation, endpoint, and distance
# ------------------------------------------------------------

p95_hq_check <- hq_summary_allozone %>%
  filter(
    !chemical %in% c(
      "Ozone (time-matched)",
      "Ozone (p50)"
    )
  ) %>%
  group_by(
    Operation,
    Endpoint,
    distance_ft
  ) %>%
  summarise(
    HI_from_p95_HQ_summary = sum(
      HQ_mean,
      na.rm = TRUE
    ),

    .groups = "drop"
  ) %>%
  left_join(
    hi_final %>%
      group_by(
        Operation,
        Endpoint,
        distance_ft
      ) %>%
      summarise(
        HI_total_p95_mean = mean(
          HI_total_p95,
          na.rm = TRUE
        ),

        .groups = "drop"
      ),

    by = c(
      "Operation",
      "Endpoint",
      "distance_ft"
    )
  ) %>%
  mutate(
    difference = HI_from_p95_HQ_summary -
      HI_total_p95_mean
  )


summary(
  p95_hq_check$difference
)

# The differences should be zero or extremely close to zero.


# ------------------------------------------------------------
# 14. Quality-control check for all three ozone scenarios
#
# The chemical component is the same for all scenarios.
# Only the selected ozone row changes.
# ------------------------------------------------------------

chemical_mean_for_scenario_check <- hq_summary_allozone %>%
  filter(
    !chemical %in% c(
      "Ozone (time-matched)",
      "Ozone (p50)",
      "Ozone (p95)"
    )
  ) %>%
  group_by(
    Operation,
    Endpoint,
    distance_ft
  ) %>%
  summarise(
    HI_chemicals_from_summary = sum(
      HQ_mean,
      na.rm = TRUE
    ),

    .groups = "drop"
  )


ozone_mean_for_scenario_check <- hq_summary_allozone %>%
  filter(
    chemical %in% c(
      "Ozone (time-matched)",
      "Ozone (p50)",
      "Ozone (p95)"
    )
  ) %>%
  select(
    Operation,
    Endpoint,
    distance_ft,
    chemical,
    ozone_HQ_mean = HQ_mean
  ) %>%
  mutate(
    HI_column = recode(
      chemical,
      "Ozone (time-matched)" = "HI_total_time",
      "Ozone (p50)" = "HI_total_p50",
      "Ozone (p95)" = "HI_total_p95"
    )
  )


hi_mean_for_scenario_check <- hi_final %>%
  group_by(
    Operation,
    Endpoint,
    distance_ft
  ) %>%
  summarise(
    HI_total_time = mean(
      HI_total_time,
      na.rm = TRUE
    ),

    HI_total_p50 = mean(
      HI_total_p50,
      na.rm = TRUE
    ),

    HI_total_p95 = mean(
      HI_total_p95,
      na.rm = TRUE
    ),

    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(
      HI_total_time,
      HI_total_p50,
      HI_total_p95
    ),

    names_to = "HI_column",
    values_to = "HI_actual_mean"
  )


allozone_hq_check <- ozone_mean_for_scenario_check %>%
  left_join(
    chemical_mean_for_scenario_check,

    by = c(
      "Operation",
      "Endpoint",
      "distance_ft"
    )
  ) %>%
  mutate(
    HI_from_HQ_summary =
      HI_chemicals_from_summary + ozone_HQ_mean
  ) %>%
  left_join(
    hi_mean_for_scenario_check,

    by = c(
      "Operation",
      "Endpoint",
      "distance_ft",
      "HI_column"
    )
  ) %>%
  mutate(
    difference =
      HI_from_HQ_summary - HI_actual_mean
  )


# Overall check
summary(
  allozone_hq_check$difference
)


# Check separately by ozone scenario
allozone_hq_check %>%
  group_by(
    chemical
  ) %>%
  summarise(
    maximum_absolute_difference = max(
      abs(difference),
      na.rm = TRUE
    ),

    mean_absolute_difference = mean(
      abs(difference),
      na.rm = TRUE
    ),

    .groups = "drop"
  )


# ------------------------------------------------------------
# 15. Save the separate all-ozone HQ summary
#
# This does not overwrite HQ_summary_wbkg.csv.
# ------------------------------------------------------------

write.csv(
  hq_summary_allozone,

  file = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute",
    "HQ_summary_wbkg_allozone.csv"
  ),

  row.names = FALSE
)

```




#### make new line graphs for MS ####




#making the same color and shapes per endpoint for all graphs
```{r}
endpoint_levels_all <- hi_final %>%
  filter(!is.na(Endpoint)) %>%
  distinct(Endpoint) %>%
  pull(Endpoint) %>%
  sort()

endpoint_palette <- c(
  "#0072B2", # blue
  "#D55E00", # vermillion
  "#009E73", # bluish green
  "#E69F00", # orange
  "#CC79A7", # reddish purple
  "#56B4E9", # sky blue
  "#F0E442", # yellow
  "#999999", # gray
  "#332288",
  "#88CCEE",
  "#44AA99",
  "#117733",
  "#DDCC77",
  "#CC6677"
)

endpoint_shape_values <- c(
  16, 15, 17, 18, 25, 22, 8, 24, 6, 3, 4, 7, 9, 10
)

endpoint_colors <- rep(endpoint_palette, length.out = length(endpoint_levels_all))
names(endpoint_colors) <- endpoint_levels_all

endpoint_shapes <- rep(endpoint_shape_values, length.out = length(endpoint_levels_all))
names(endpoint_shapes) <- endpoint_levels_all
```



#individual plots per operation, line graphs colored by endpoint, avg HI vs. distance, ozone scenario

#making 4 plots: 1) p95 with all endpoints, 2) p50 with all endpoints, 3) p95 with exceeding endpoints, 4) p50 with exceeding endpoints
```{r}
# ============================================================
# Helper function for operation-specific line plots
#
# Creates four plot sets:
# 1. p95 ozone, all endpoints
# 2. p50 ozone, all endpoints
# 3. p95 ozone, endpoints exceeding average HI > 1
# 4. p50 ozone, endpoints exceeding average HI > 1

# ============================================================

make_operation_hi_plots <- function(
  hi_col,
  ozone_label,
  exceed_only = FALSE,
  output_prefix
) {

  hi_col_sym <- rlang::sym(hi_col)


  # ----------------------------------------------------------
  # 1. Summarize HI by operation, endpoint, and distance
  # ----------------------------------------------------------

  hi_summary <- hi_final %>%
    filter(
      !is.na(Operation),
      !is.na(Endpoint),
      !is.na(!!hi_col_sym)
    ) %>%
    group_by(
      Operation,
      Endpoint,
      distance_ft
    ) %>%
    summarise(
      HI_mean = mean(
        !!hi_col_sym,
        na.rm = TRUE
      ),

      HI_median = median(
        !!hi_col_sym,
        na.rm = TRUE
      ),

      HI_sd = sd(
        !!hi_col_sym,
        na.rm = TRUE
      ),

      n = sum(
        !is.na(!!hi_col_sym)
      ),

      HI_se = HI_sd / sqrt(n),

      HI_min = min(
        !!hi_col_sym,
        na.rm = TRUE
      ),

      HI_max = max(
        !!hi_col_sym,
        na.rm = TRUE
      ),

      .groups = "drop"
    ) %>%

    mutate(
      HI_ymin = pmax(
        HI_mean - HI_se,
        0
      ),

      HI_ymax = HI_mean + HI_se,

      Endpoint = factor(
        Endpoint,
        levels = endpoint_levels_all
      )
    )


  # ----------------------------------------------------------
  # 2. Calculate sample size for each operation
  # ----------------------------------------------------------

  operation_n <- hi_final %>%
    filter(
      !is.na(Operation),
      !is.na(Endpoint),
      !is.na(!!hi_col_sym)
    ) %>%

    distinct(
      Operation,
      `Start Date Time`,
      `Source Wellpad`,
      `Sampling Site`
    ) %>%

    count(
      Operation,
      name = "n_samples"
    )


  operation_n_lookup <- setNames(
    operation_n$n_samples,
    operation_n$Operation
  )


  # ----------------------------------------------------------
  # 3. Optionally retain endpoints exceeding average HI > 1
  #    within each operation
  # ----------------------------------------------------------

  if (exceed_only) {

    operation_endpoint_exceeding <- hi_summary %>%
      group_by(
        Operation,
        Endpoint
      ) %>%

      summarise(
        HI_max_operation = max(
          HI_mean,
          na.rm = TRUE
        ),
        .groups = "drop"
      ) %>%

      filter(
        HI_max_operation > 1
      )


    hi_summary <- hi_summary %>%
      semi_join(
        operation_endpoint_exceeding,
        by = c(
          "Operation",
          "Endpoint"
        )
      )
  }


  # ----------------------------------------------------------
  # 4. Save summarized data
  # ----------------------------------------------------------

  write.csv(
    hi_summary,

    file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/acute/Line_Operation",

      paste0(
        "HI_summary_",
        output_prefix,
        "_wbkg.csv"
      )
    ),

    row.names = FALSE
  )


  # ----------------------------------------------------------
  # 5. Define operation order
  # ----------------------------------------------------------

  operation_names <- sort(
    unique(
      hi_final$Operation[
        !is.na(hi_final$Operation)
      ]
    )
  )


  n_operations <- length(
    operation_names
  )


  n_panel_columns <- 2


  # Number of rows after adding the legend placeholder
  n_total_slots <- n_operations + 1


  n_panel_rows <- ceiling(
    n_total_slots / n_panel_columns
  )


  # Equal-height rows, matching the HQ bar plot structure
  row_heights <- rep(
    1,
    n_panel_rows
  )


  # ----------------------------------------------------------
  # 6. Define endpoints used in the shared legend
  # ----------------------------------------------------------

  if (exceed_only) {

    # Preferred operation from which to construct the
    # shared exceedance legend
    legend_operation <- "Coiled Tubing"


    # Use the first operation with data as a fallback
    if (
      !legend_operation %in%
        unique(hi_summary$Operation)
    ) {

      legend_operation <- hi_summary %>%
        filter(
          !is.na(Operation)
        ) %>%

        distinct(
          Operation
        ) %>%

        pull(
          Operation
        ) %>%

        first()
    }


    endpoint_levels_legend <- hi_summary %>%
      filter(
        Operation == legend_operation
      ) %>%

      distinct(
        Endpoint
      ) %>%

      pull(
        Endpoint
      ) %>%

      as.character()


    endpoint_levels_legend <- endpoint_levels_all[
      endpoint_levels_all %in%
        endpoint_levels_legend
    ]

  } else {

    endpoint_levels_legend <- endpoint_levels_all
  }


  # ----------------------------------------------------------
  # 7. Create one plot per operation
  # ----------------------------------------------------------

  plots_by_operation <- lapply(
    seq_along(operation_names),

    function(op_index) {

      op <- operation_names[op_index]

      op_n <- operation_n_lookup[[op]]


      operation_df <- hi_summary %>%
        filter(
          Operation == op
        ) %>%

        droplevels()


      # Endpoints actually present in this operation
      operation_endpoint_levels <- operation_df %>%
        distinct(
          Endpoint
        ) %>%

        pull(
          Endpoint
        ) %>%

        as.character()


      # Preserve global endpoint ordering
      operation_endpoint_levels <- endpoint_levels_all[
        endpoint_levels_all %in%
          operation_endpoint_levels
      ]


      # Operation-specific x-axis breaks
      x_breaks <- sort(
        unique(
          operation_df$distance_ft
        )
      )


      # Thin breaks only if there are substantially more than
      # the standard 500-foot intervals
      if (length(x_breaks) > 12) {

        x_breaks_display <- x_breaks[
          seq(
            from = 1,
            to = length(x_breaks),
            by = 2
          )
        ]

      } else {

        x_breaks_display <- x_breaks
      }


      p <- ggplot(
        operation_df,

        aes(
          x = distance_ft,
          y = HI_mean,
          color = Endpoint,
          shape = Endpoint,
          group = Endpoint
        )
      ) +

        geom_line(
          linewidth = 1
        ) +

        geom_point(
          size = 2.3,
          stroke = 0.8
        ) +

        geom_errorbar(
          aes(
            ymin = HI_ymin,
            ymax = HI_ymax
          ),
          width = 50,
          alpha = 0.45,
          linewidth = 0.35
        ) +

        geom_hline(
          yintercept = 1,
          linetype = "dashed",
          linewidth = 0.6
        ) +

        geom_vline(
          xintercept = 2000,
          linetype = "dotted",
          color = "red",
          linewidth = 0.7
        ) +

        scale_color_manual(
          name = "Endpoint",
          values = endpoint_colors,
          breaks = operation_endpoint_levels,
          limits = operation_endpoint_levels,
          drop = TRUE
        ) +

        scale_shape_manual(
          name = "Endpoint",
          values = endpoint_shapes,
          breaks = operation_endpoint_levels,
          limits = operation_endpoint_levels,
          drop = TRUE
        ) +

        scale_x_continuous(
          breaks = x_breaks_display,

          expand = expansion(
            mult = c(
              0.02,
              0.02
            )
          )
        ) +

        scale_y_continuous(
          expand = expansion(
            mult = c(
              0,
              0.04
            )
          )
        ) +

labs(
  tag = LETTERS[op_index],
  title = NULL,
  x = NULL,
  y = NULL
) +

        guides(
          color = guide_legend(
            ncol = 1,
            byrow = TRUE,

            override.aes = list(
              linewidth = 1.1,
              size = 3.5
            )
          ),

          shape = guide_legend(
            ncol = 1,
            byrow = TRUE,

            override.aes = list(
              size = 3.5
            )
          )
        ) +

        theme_minimal(
          base_size = 11
        ) +

        theme(
          plot.title = element_text(
            hjust = 0.5,
            face = "bold",
            size = 12,
            lineheight = 0.95,

            margin = margin(
              b = 6
            )
          ),

          # Panel-label appearance and position
          plot.tag = element_text(
            face = "bold",
            size = 11,
            hjust = 0,
            vjust = 1
          ),

plot.tag.position = c(
  0,
  1.06
),

          # Retain x-axis labels and ticks on every panel
          axis.text.x = element_text(
            angle = 45,
            hjust = 1,
            vjust = 1,
            size = 9.5
          ),

          axis.ticks.x = element_line(
            linewidth = 0.3
          ),

          axis.text.y = element_text(
            size = 10
          ),

          axis.ticks.y = element_line(
            linewidth = 0.3
          ),

          # Remove individual panel axis titles
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),

          legend.title = element_text(
            face = "bold",
            size = 11
          ),

          legend.text = element_text(
            size = 10
          ),

          legend.key.size = grid::unit(
            0.44,
            "cm"
          ),

          legend.key.height = grid::unit(
            0.44,
            "cm"
          ),

          legend.key.width = grid::unit(
            0.44,
            "cm"
          ),

          legend.spacing.y = grid::unit(
            0.03,
            "cm"
          ),

          legend.margin = margin(
            t = 0,
            r = 0,
            b = 0,
            l = 0
          ),

          panel.grid.minor = element_blank(),

          # Slightly more space for panel tag
plot.margin = margin(
  t = 14,
  r = 5,
  b = 5,
  l = 8
)
        )


      p
    }
  )


  names(
    plots_by_operation
  ) <- operation_names


  # ----------------------------------------------------------
  # 8. Create combined two-column graph section
  # ----------------------------------------------------------

  if (exceed_only) {

    # --------------------------------------------------------
    # Exceedance plots:
    # create a dedicated legend and place it in the final slot
    # --------------------------------------------------------

    legend_source_plot <-
      plots_by_operation[[legend_operation]] +

      scale_color_manual(
        name = "Endpoint",
        values = endpoint_colors,
        breaks = endpoint_levels_legend,
        limits = endpoint_levels_legend,
        drop = FALSE
      ) +

      scale_shape_manual(
        name = "Endpoint",
        values = endpoint_shapes,
        breaks = endpoint_levels_legend,
        limits = endpoint_levels_legend,
        drop = FALSE
      ) +

      guides(
        color = guide_legend(
          ncol = 1,
          byrow = TRUE,

          override.aes = list(
            linewidth = 1.1,
            size = 3.5
          )
        ),

        shape = guide_legend(
          ncol = 1,
          byrow = TRUE,

          override.aes = list(
            size = 3.5
          )
        )
      ) +

      theme(
        legend.position = "right",
        legend.justification = "center",
        legend.direction = "vertical",
        legend.box = "vertical",

        legend.title = element_text(
          face = "bold",
          size = 11
        ),

        legend.text = element_text(
          size = 10
        ),

        legend.key.size = grid::unit(
          0.44,
          "cm"
        ),

        legend.key.height = grid::unit(
          0.44,
          "cm"
        ),

        legend.key.width = grid::unit(
          0.44,
          "cm"
        ),

        legend.spacing.y = grid::unit(
          0.03,
          "cm"
        ),

        legend.margin = margin(
          t = 0,
          r = 0,
          b = 0,
          l = 0
        )
      )


    legend_grob <- cowplot::get_legend(
      legend_source_plot
    )


    legend_panel <- patchwork::wrap_elements(
      full = cowplot::ggdraw(
        legend_grob
      ),
      ignore_tag = TRUE
    )


    plots_without_legends <- lapply(
      plots_by_operation,

      function(p) {

        p +
          theme(
            legend.position = "none"
          )
      }
    )


    plot_list_with_legend <- c(
      plots_without_legends,

      list(
        legend_panel
      )
    )


    graph_section <- patchwork::wrap_plots(
      plotlist = plot_list_with_legend,
      ncol = 2,
      heights = row_heights
    )

  } else {

    # --------------------------------------------------------
    # All-endpoint plots:
    # use the complete shared endpoint legend
    # --------------------------------------------------------

    panel_plots <- lapply(
      plots_by_operation,

      function(p) {

        p +

          scale_color_manual(
            name = "Endpoint",
            values = endpoint_colors,
            breaks = endpoint_levels_all,
            limits = endpoint_levels_all,
            drop = FALSE
          ) +

          scale_shape_manual(
            name = "Endpoint",
            values = endpoint_shapes,
            breaks = endpoint_levels_all,
            limits = endpoint_levels_all,
            drop = FALSE
          ) +

          guides(
            color = guide_legend(
              ncol = 1,
              byrow = TRUE,

              override.aes = list(
                linewidth = 1.1,
                size = 3.5
              )
            ),

            shape = guide_legend(
              ncol = 1,
              byrow = TRUE,

              override.aes = list(
                size = 3.5
              )
            )
          )
      }
    )


    plot_list_with_legend <- c(
      panel_plots,

      list(
        patchwork::guide_area()
      )
    )


    graph_section <- patchwork::wrap_plots(
      plotlist = plot_list_with_legend,
      ncol = 2,
      guides = "collect",
      heights = row_heights
    ) &

      theme(
        legend.position = "right",
        legend.justification = "center",
        legend.direction = "vertical",
        legend.box = "vertical",

        legend.title = element_text(
          face = "bold",
          size = 11
        ),

        legend.text = element_text(
          size = 10
        ),

        legend.key.size = grid::unit(
          0.44,
          "cm"
        ),

        legend.key.height = grid::unit(
          0.44,
          "cm"
        ),

        legend.key.width = grid::unit(
          0.44,
          "cm"
        ),

        legend.spacing.y = grid::unit(
          0.03,
          "cm"
        ),

        legend.margin = margin(
          t = 0,
          r = 0,
          b = 0,
          l = 0
        )
      )
  }


  # ----------------------------------------------------------
  # 9. Create one shared x-axis title
  # ----------------------------------------------------------

  shared_x_axis_title <- patchwork::wrap_elements(
    full = grid::textGrob(
      label = "Distance from Well Pad (ft)",
      x = 0.5,
      y = 0.65,
      just = "center",

      gp = grid::gpar(
        fontsize = 11.5
      )
    ),

    ignore_tag = TRUE
  )


  # ----------------------------------------------------------
  # 10. Create one shared y-axis title
  # ----------------------------------------------------------

  shared_y_axis_title <- patchwork::wrap_elements(
    full = grid::textGrob(
      label = paste0(
        "Average HI (",
        ozone_label,
        " ozone)"
      ),
      x = 0.55,
      y = 0.5,
      just = "center",
      rot = 90,

      gp = grid::gpar(
        fontsize = 11.5
      )
    ),

    ignore_tag = TRUE
  )


  # ----------------------------------------------------------
  # 11. Add shared y-axis title beside graph section
  # ----------------------------------------------------------

  graph_with_y_title <- (
    shared_y_axis_title |
      graph_section
  ) +

    patchwork::plot_layout(
      widths = c(
        0.045,
        1
      )
    )


  # ----------------------------------------------------------
  # 12. Add shared x-axis title beneath complete figure
  # ----------------------------------------------------------

  combined_plot <- (
    graph_with_y_title /
      shared_x_axis_title
  ) +

    patchwork::plot_layout(
      heights = c(
        1,
        0.035
      )
    )


  # ----------------------------------------------------------
  # 13. Save combined panel plot
  # ----------------------------------------------------------

  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/acute/Line_Operation",

      paste0(
        "HI_",
        output_prefix,
        "_wbkg_panel.png"
      )
    ),

    plot = combined_plot,

    # Match the dimensions and resolution of the HQ bar plots
    width = 7,
    height = 10,
    units = "in",
    dpi = 400,
    bg = "white"
  )


  # ----------------------------------------------------------
  # 14. Return results
  # ----------------------------------------------------------

  return(
    list(
      summary = hi_summary,
      plots = plots_by_operation,
      graph_section = graph_section,
      panel = combined_plot
    )
  )
}



# ============================================================
# Run all four plot sets
# ============================================================

p95_all <- make_operation_hi_plots(
  hi_col = "HI_total_p95",
  ozone_label = "p95",
  exceed_only = FALSE,
  output_prefix = "p95"
)


p50_all <- make_operation_hi_plots(
  hi_col = "HI_total_p50",
  ozone_label = "p50",
  exceed_only = FALSE,
  output_prefix = "p50"
)


p95_exceed <- make_operation_hi_plots(
  hi_col = "HI_total_p95",
  ozone_label = "p95",
  exceed_only = TRUE,
  output_prefix = "p95_exceedHI1"
)


p50_exceed <- make_operation_hi_plots(
  hi_col = "HI_total_p50",
  ozone_label = "p50",
  exceed_only = TRUE,
  output_prefix = "p50_exceedHI1"
)



# ============================================================
# View final combined panels
# ============================================================

p95_all$panel

p50_all$panel

p95_exceed$panel

p50_exceed$panel
```




#pulling data on just the min, max, average for each endpoint and operation at every distance in the line graphs above for the MS (p95 ozone)
```{r}
#long format
hi_exceedance_table <- p95_exceed$summary %>%
  select(
    Operation,
    Endpoint,
    distance_ft,
    n,
    HI_mean,
    HI_min,
    HI_max
  ) %>%
  arrange(
    Operation,
    Endpoint,
    distance_ft
  )


#wide format
hi_publication_table_wide <- p95_exceed$summary %>%
  select(
    Operation,
    Endpoint,
    distance_ft,
    n,
    HI_mean,
    HI_min,
    HI_max
  ) %>%

  # Add "ft" to the distance portion of each column name
  mutate(
    distance_label = paste0(
      distance_ft,
      "ft"
    )
  ) %>%

  select(
    Operation,
    Endpoint,
    n,
    distance_label,
    HI_mean,
    HI_min,
    HI_max
  ) %>%

  pivot_wider(
    names_from = distance_label,
    values_from = c(
      HI_mean,
      HI_min,
      HI_max
    ),
    names_glue = "{.value}_{distance_label}"
  ) %>%

  arrange(
    Operation,
    Endpoint
  )


# Save the table
write.csv(
  hi_publication_table_wide,

  file = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute/Line_Operation",
    "Table_HI_p95_exceedHI1_wide_by_distance.csv"
  ),

  row.names = FALSE
)
```





#### graphs for just the one separator maintenance scenario ####


#The separator maintenance trigger canister is the one dated 6/23/2021 from the Anthem sensor (labeled as production)

#plot of total HI vs. distance
```{r}
# ------------------------------------------------------------
# 1. Define event time
# ------------------------------------------------------------

event_time <- as.POSIXct(
  "2021-06-23 09:48",
  tz = "UTC"
)


# ------------------------------------------------------------
# 2. Filter to the selected event
# ------------------------------------------------------------

event_df <- hi_final %>%
  filter(
    `Start Date Time` == event_time
  )


# ------------------------------------------------------------
# 3. Identify endpoints with HI > 1 specifically at 500 ft
# ------------------------------------------------------------

endpoints_exceeding_500ft <- event_df %>%
  filter(
    distance_ft == 500,
    !is.na(HI_total_p95)
  ) %>%
  group_by(
    Endpoint
  ) %>%
  summarise(
    HI_500ft = max(
      HI_total_p95,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  filter(
    HI_500ft > 1
  ) %>%
  pull(
    Endpoint
  ) %>%
  as.character()


# ------------------------------------------------------------
# 4. Preserve the global endpoint order
# ------------------------------------------------------------

endpoint_levels_plot <- endpoint_levels_all[
  endpoint_levels_all %in%
    endpoints_exceeding_500ft
]


# ------------------------------------------------------------
# 5. Filter plotting dataset
# ------------------------------------------------------------

plot_df <- event_df %>%
  filter(
    Endpoint %in% endpoint_levels_plot
  ) %>%
  mutate(
    Endpoint = factor(
      Endpoint,
      levels = endpoint_levels_plot
    )
  )


# ------------------------------------------------------------
# 6. Confirm that colors and shapes are being matched by name
# ------------------------------------------------------------

endpoint_style_check <- data.frame(
  Endpoint = endpoint_levels_plot,
  color = unname(
    endpoint_colors[
      endpoint_levels_plot
    ]
  ),
  shape = unname(
    endpoint_shapes[
      endpoint_levels_plot
    ]
  )
)

print(
  endpoint_style_check
)


# ------------------------------------------------------------
# 7. Define x-axis breaks every 500 ft
# ------------------------------------------------------------

x_min <- floor(
  min(
    plot_df$distance_ft,
    na.rm = TRUE
  ) / 500
) * 500


x_max <- ceiling(
  max(
    plot_df$distance_ft,
    na.rm = TRUE
  ) / 500
) * 500


x_breaks_500 <- seq(
  from = x_min,
  to = x_max,
  by = 500
)


# ------------------------------------------------------------
# 8. Create plot
# ------------------------------------------------------------

hi_distance_ex <- ggplot(
  plot_df,
  aes(
    x = distance_ft,
    y = HI_total_p95,
    color = Endpoint,
    shape = Endpoint,
    group = Endpoint
  )
) +

  geom_line(
    linewidth = 1
  ) +

  geom_point(
    size = 2.3,
    stroke = 0.8
  ) +

  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    linewidth = 0.6
  ) +
  
  geom_vline(
    xintercept = 2000,
    linetype = "dotted",
    color = "red",
    linewidth = 0.8
  ) +

  scale_color_manual(
    name = "Endpoint",

    # Subset the original named vector by endpoint name
    values = endpoint_colors[
      endpoint_levels_plot
    ],

    breaks = endpoint_levels_plot,
    limits = endpoint_levels_plot,
    drop = TRUE
  ) +

  scale_shape_manual(
    name = "Endpoint",

    # Subset the original named vector by endpoint name
    values = endpoint_shapes[
      endpoint_levels_plot
    ],

    breaks = endpoint_levels_plot,
    limits = endpoint_levels_plot,
    drop = TRUE
  ) +

  scale_x_continuous(
    breaks = x_breaks_500,
    limits = c(
      x_min,
      x_max
    ),
    expand = expansion(
      mult = c(
        0.01,
        0.01
      )
    )
  ) +

  scale_y_continuous(
    expand = expansion(
      mult = c(
        0,
        0.04
      )
    )
  ) +

  labs(
    x = "Distance from Well Pad (ft)",
    y = "Hazard Index (HI) - p95 Ozone",
    color = "Endpoint",
    shape = "Endpoint",
    title = NULL
  ) +

  guides(
    color = guide_legend(
      ncol = 1,
      byrow = TRUE,
      override.aes = list(
        linewidth = 1.1,
        size = 3.5
      )
    ),

    shape = guide_legend(
      ncol = 1,
      byrow = TRUE,
      override.aes = list(
        size = 3.5
      )
    )
  ) +

  theme_minimal(
    base_size = 11
  ) +

theme(
  axis.text.x = element_text(
    angle = 45,
    hjust = 1,
    vjust = 1,
    size = 9.5
  ),

  axis.text.y = element_text(
    size = 10
  ),

  axis.title.x = element_text(
    size = 11.5,
    margin = margin(t = 8)
  ),

  axis.title.y = element_text(
    size = 11.5,
    margin = margin(r = 8)
  ),

  legend.position = c(0.65, 0.78),
  legend.justification = c(0, 1),

  # Give the legend a white box
  legend.background = element_rect(
    fill = "white",
    colour = "black",
    linewidth = 0.3
  ),

  legend.box.background = element_blank(),

  legend.title = element_text(
    face = "bold",
    size = 11
  ),

  legend.text = element_text(
    size = 10
  ),

  legend.key.size = grid::unit(
    0.44,
    "cm"
  ),

  legend.key.height = grid::unit(
    0.44,
    "cm"
  ),

  legend.key.width = grid::unit(
    0.44,
    "cm"
  ),

  legend.spacing.y = grid::unit(
    0.03,
    "cm"
  ),

  panel.grid.minor = element_blank()
)


# ------------------------------------------------------------
# 9. View plot
# ------------------------------------------------------------

hi_distance_ex


# ------------------------------------------------------------
# 10. Save plot
# ------------------------------------------------------------

ggsave(
  filename = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute",
    "sepmain_2021-06-23_HIvdistance.png"
  ),

  plot = hi_distance_ex,

  width = 6,
  height = 4,
  units = "in",
  dpi = 300,
  bg = "white"
)


# ------------------------------------------------------------
# 11. Save plotted data
# ------------------------------------------------------------

write.csv(
  plot_df,

  file = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute",
    "sepmain_2021-06-23_HIvdistance.csv"
  ),

  row.names = FALSE
)

# ------------------------------------------------------------
# 11. Save data for all endpoints for the selected event
# ------------------------------------------------------------

event_df_all_endpoints <- event_df %>%
  mutate(
    Endpoint = factor(
      Endpoint,
      levels = endpoint_levels_all
    )
  ) %>%
  arrange(
    Endpoint,
    distance_ft
  )


write.csv(
  event_df_all_endpoints,

  file = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute",
    "sepmain_2021-06-23_HIvdistance_all_endpoints.csv"
  ),

  row.names = FALSE
)
```


#plot of chemical contributions
```{r}
# ------------------------------------------------------------
# 1. Filter HQ data for the selected event and endpoints
# ------------------------------------------------------------

hq_event_df <- hq_all %>%
  filter(
    `Start Date Time` == event_time
  ) %>%
  group_by(
    Operation,
    Endpoint,
    distance_ft,
    chemical
  ) %>%
  summarise(
    HQ_mean = mean(
      HQ,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  filter(
    Endpoint %in% endpoint_levels_plot
  )


# ------------------------------------------------------------
# 2. Clean chemical and endpoint names
# ------------------------------------------------------------

hq_event_df <- hq_event_df %>%
  mutate(
    chemical = recode(
      chemical,
      "aliphatic_low"   = "Aliphatic low",
      "aromatic_medium" = "Aromatic medium",
      "benzene"         = "Benzene",
      "c2cl4"           = "PCE",
      "c2hcl3"          = "TCE",
      "ethylbenzene"    = "Ethylbenzene",
      "n_nonane"        = "n_Nonane",
      "ozone"           = "Ozone (p95)",
      "styrene"         = "Styrene",
      "toluene"         = "Toluene",
      "xylenes"         = "Xylenes"
    ),

    Endpoint_label = stringr::str_to_title(
      as.character(Endpoint)
    )
  )


# ------------------------------------------------------------
# 3. Define chemical legend labels
# ------------------------------------------------------------

chemical_legend_labels <- c(
  "Aliphatic low"   = "Aliphatic low (Al)",
  "Aromatic medium" = "Aromatic medium (Am)",
  "Benzene"         = "Benzene (B)",
  "Ethylbenzene"    = "Ethylbenzene (E)",
  "n_Nonane"        = "n_Nonane (N)",
  "Ozone (p95)"     = "p95 Ozone (O)",
  "PCE"             = "PCE (P)",
  "Styrene"         = "Styrene (S)",
  "TCE"             = "TCE (T)",
  "Toluene"         = "Toluene (To)",
  "Xylenes"         = "Xylenes (X)"
)


# ------------------------------------------------------------
# 4. Define the six endpoints to be plotted
# ------------------------------------------------------------

endpoint_names <- hq_event_df %>%
  distinct(
    Endpoint,
    Endpoint_label
  ) %>%
  mutate(
    Endpoint_order = match(
      Endpoint,
      endpoint_levels_plot
    )
  ) %>%
  arrange(
    Endpoint_order
  ) %>%
  pull(
    Endpoint
  ) %>%
  as.character()


# Restrict to the six endpoint graphs being shown
endpoint_names <- endpoint_names[
  seq_len(
    min(
      6,
      length(endpoint_names)
    )
  )
]


# ------------------------------------------------------------
# 5. Retain only chemicals that contribute HQ > 0 in at least
#    one of the six endpoint graphs
# ------------------------------------------------------------

chemicals_present <- hq_event_df %>%
  filter(
    Endpoint %in% endpoint_names,
    !is.na(chemical),
    !is.na(HQ_mean),
    HQ_mean > 0
  ) %>%
  distinct(
    chemical
  ) %>%
  pull(
    chemical
  ) %>%
  as.character()


# Preserve the order defined in chemical_colors
chemical_order <- names(chemical_colors)[
  names(chemical_colors) %in% chemicals_present
]


# Confirm the result
print(
  chemical_order
)


# Optional table showing each chemical's contribution across
# the six displayed endpoint graphs
chemical_contribution_check <- hq_event_df %>%
  filter(
    Endpoint %in% endpoint_names,
    !is.na(chemical)
  ) %>%
  group_by(
    chemical
  ) %>%
  summarise(
    max_HQ = max(
      HQ_mean,
      na.rm = TRUE
    ),

    total_HQ = sum(
      HQ_mean,
      na.rm = TRUE
    ),

    endpoints_contributed_to = n_distinct(
      Endpoint[
        HQ_mean > 0
      ]
    ),

    .groups = "drop"
  ) %>%
  arrange(
    desc(total_HQ)
  )


print(
  chemical_contribution_check
)


# Create the final legend labels for only those chemicals
chemical_labels_used <- chemical_legend_labels[
  chemical_order
]


# Apply the same reduced chemical order to the plotting data
hq_event_df <- hq_event_df %>%
  filter(
    Endpoint %in% endpoint_names
  ) %>%
  mutate(
    chemical = factor(
      chemical,
      levels = chemical_order
    )
  )


# ------------------------------------------------------------
# 6. Create one plot per endpoint
# ------------------------------------------------------------

endpoint_plots <- lapply(
  seq_along(endpoint_names),
  function(endpoint_index) {

    ep <- endpoint_names[endpoint_index]

    endpoint_df <- hq_event_df %>%
      filter(
        Endpoint == ep
      ) %>%
      mutate(
        # Keep the same chemical levels in every graph
        chemical = factor(
          as.character(chemical),
          levels = chemical_order
        )
      )

    endpoint_title <- endpoint_df %>%
      distinct(Endpoint_label) %>%
      pull(Endpoint_label) %>%
      first()

    # Sort distances numerically
    distance_levels <- endpoint_df %>%
      distinct(distance_ft) %>%
      arrange(distance_ft) %>%
      pull(distance_ft) %>%
      as.character()

    endpoint_df <- endpoint_df %>%
      mutate(
        distance_factor = factor(
          as.character(distance_ft),
          levels = distance_levels
        )
      )

    ggplot(
      endpoint_df,
      aes(
        x = distance_factor,
        y = HQ_mean,
        fill = chemical
      )
    ) +

      geom_col(
        position = "stack",
        width = 0.8
      ) +

      geom_hline(
        yintercept = 1,
        linetype = "dashed",
        linewidth = 0.6
      ) +

      scale_fill_manual(
        name = "Chemical",
        values = chemical_colors[chemical_order],
        breaks = chemical_order,
        limits = chemical_order,
        labels = chemical_labels_used,

        # Keep the identical legend definition in every graph
        drop = FALSE
      ) +

      scale_y_continuous(
        expand = expansion(
          mult = c(0, 0.03)
        )
      ) +

      labs(
        tag = LETTERS[endpoint_index],
        title = NULL,
        x = NULL,
        y = NULL
      ) +

      guides(
        fill = guide_legend(
          ncol = 4,
          byrow = TRUE
        )
      ) +

      theme_minimal(
        base_size = 11
      ) +

      theme(
        plot.title = element_text(
          hjust = 0.5,
          face = "bold",
          size = 12,
          lineheight = 0.95,
          margin = margin(b = 6)
        ),

plot.tag = element_text(
  face = "bold",
  size = 11,
  hjust = 0,
  vjust = 0
),

plot.tag.position = c(
  -0.06,
  1.00
),

        axis.text.x = element_text(
          angle = 45,
          hjust = 1,
          vjust = 1,
          size = 9.5
        ),

        axis.text.y = element_text(
          size = 10
        ),

        axis.ticks.x = element_line(
          linewidth = 0.3
        ),

        axis.ticks.y = element_line(
          linewidth = 0.3
        ),

        axis.title.x = element_blank(),
        axis.title.y = element_blank(),

        legend.title = element_text(
          face = "bold",
          size = 11
        ),

        legend.text = element_text(
          size = 9.5
        ),

        legend.key.size = grid::unit(
          0.4,
          "cm"
        ),

        legend.key.height = grid::unit(
          0.4,
          "cm"
        ),

        legend.key.width = grid::unit(
          0.4,
          "cm"
        ),

        legend.spacing.x = grid::unit(
          0.15,
          "cm"
        ),

        legend.spacing.y = grid::unit(
          0.08,
          "cm"
        ),

        legend.margin = margin(
          t = 2,
          r = 2,
          b = 2,
          l = 2
        ),

        panel.grid.minor = element_blank(),

        plot.margin = margin(
          t = 8,
          r = 5,
          b = 5,
          l = 8
        )
      )
  }
)

names(endpoint_plots) <- endpoint_names

# ------------------------------------------------------------
# 7. Create one dedicated shared legend containing only
#    chemicals with HQ > 0 in at least one displayed panel
# ------------------------------------------------------------

legend_df <- data.frame(
  chemical = factor(
    chemical_order,
    levels = chemical_order
  ),

  x = seq_along(
    chemical_order
  ),

  y = 1
)


legend_source_plot <- ggplot(
  legend_df,
  aes(
    x = x,
    y = y,
    fill = chemical
  )
) +

  geom_col() +

  scale_fill_manual(
    name = "Chemical",
    values = chemical_colors[
      chemical_order
    ],
    breaks = chemical_order,
    limits = chemical_order,
    labels = chemical_labels_used,
    drop = FALSE
  ) +

  guides(
    fill = guide_legend(
      ncol = 4,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +

  theme_void() +

  theme(
    legend.position = "bottom",
    legend.justification = "center",
    legend.direction = "horizontal",
    legend.box = "horizontal",

    legend.title = element_text(
      face = "bold",
      size = 11,
      hjust = 0.5
    ),

    legend.text = element_text(
      size = 9.5
    ),

    legend.key.size = grid::unit(
      0.4,
      "cm"
    ),

    legend.key.height = grid::unit(
      0.4,
      "cm"
    ),

    legend.key.width = grid::unit(
      0.4,
      "cm"
    ),

    legend.spacing.x = grid::unit(
      0.15,
      "cm"
    ),

    legend.spacing.y = grid::unit(
      0.08,
      "cm"
    ),

    legend.margin = margin(
      t = 2,
      r = 2,
      b = 2,
      l = 2
    )
  )


shared_legend_grob <- cowplot::get_legend(
  legend_source_plot
)


shared_legend_panel <- patchwork::wrap_elements(
  full = shared_legend_grob,
  ignore_tag = TRUE
)


# ------------------------------------------------------------
# 8. Remove legends from all six endpoint graphs
# ------------------------------------------------------------

endpoint_plots_noleg <- lapply(
  endpoint_plots,
  function(p) {

    p +
      theme(
        legend.position = "none"
      )
  }
)


# ------------------------------------------------------------
# 9. Assemble the six endpoint graphs in two columns
# ------------------------------------------------------------

graph_section <- patchwork::wrap_plots(
  plotlist = endpoint_plots_noleg,
  ncol = 2
)


# ------------------------------------------------------------
# 10. Create the shared y-axis title
# ------------------------------------------------------------

shared_y_axis_title <- patchwork::wrap_elements(
  full = grid::textGrob(
    label = "Hazard Quotient (HQ)",
    x = 0.55,
    y = 0.5,
    just = "center",
    rot = 90,

    gp = grid::gpar(
      fontsize = 11.5
    )
  ),

  ignore_tag = TRUE
)


# ------------------------------------------------------------
# 11. Put the shared y-axis title beside the graph panels
# ------------------------------------------------------------

graph_with_y_title <- (
  shared_y_axis_title |
    graph_section
) +

  patchwork::plot_layout(
    widths = c(
      0.045,
      1
    )
  )


# ------------------------------------------------------------
# 12. Create the shared x-axis title
# ------------------------------------------------------------

shared_x_axis_title <- patchwork::wrap_elements(
  full = grid::textGrob(
    label = "Distance from Well Pad (ft)",
    x = 0.5,
    y = 0.65,
    just = "center",

    gp = grid::gpar(
      fontsize = 11.5
    )
  ),

  ignore_tag = TRUE
)


# ------------------------------------------------------------
# 13. Assemble graphs, one legend, and one x-axis title
# ------------------------------------------------------------

hq_event_plot <- (
  graph_with_y_title /
    shared_x_axis_title /
    shared_legend_panel
) +
  patchwork::plot_layout(
    heights = c(
      1,
      0.035,
      0.13
    )
  )


# ------------------------------------------------------------
# 14. View completed figure
# ------------------------------------------------------------

hq_event_plot


# ------------------------------------------------------------
# 15. Save completed figure
# ------------------------------------------------------------

ggsave(
  filename = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute",
    "sepmain_2021-06-23_HQvdistance.png"
  ),

  plot = hq_event_plot,

  width = 7,
  height = 9,
  units = "in",
  dpi = 400,
  bg = "white"
)


# ------------------------------------------------------------
# 14. Create and save raw HQ data for reporting
# ------------------------------------------------------------

hq_event_raw <- hq_all %>%
  filter(
    `Start Date Time` == event_time
  )


write.csv(
  hq_event_raw,

  file = file.path(
    "~/CO_CRA/CO_CRA_aggregate/output/acute",
    "sepmain_2021-06-23_HQs.csv"
  ),

  row.names = FALSE
)

```




#what fraction of the exposure for this event is attributable to the acute event and the monthly data
```{r}
# Definte event
event_time <- as.POSIXct("2021-06-23 09:48", tz = "UTC")

# Chronic concentrations
chronic_event <- chronic_ready %>%
  filter(`Start Date Time` == event_time) %>%
  group_by(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    chemical
  ) %>%
  summarise(
    chronic_concentration = sum(concentration, na.rm = TRUE),
    .groups = "drop"
  )

# Acute concentrations
acute_event <- acute_clean %>%
  filter(`Start Date Time` == event_time) %>%
  group_by(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    chemical
  ) %>%
  summarise(
    acute_concentration = sum(concentration, na.rm = TRUE),
    .groups = "drop"
  )

# Join together
source_contributions <- full_join(
  acute_event,
  chronic_event,
  by = c(
    "Start Date Time",
    "Source Wellpad",
    "Sampling Site",
    "distance_ft",
    "chemical"
  )
) %>%
  mutate(
    acute_concentration   = coalesce(acute_concentration, 0),
    chronic_concentration = coalesce(chronic_concentration, 0),

    total_concentration =
      acute_concentration + chronic_concentration,

    pct_acute =
      if_else(
        total_concentration > 0,
        100 * acute_concentration / total_concentration,
        NA_real_
      ),

    pct_chronic =
      if_else(
        total_concentration > 0,
        100 * chronic_concentration / total_concentration,
        NA_real_
      )
  ) %>%
  arrange(chemical, distance_ft)

# Save
write.csv(
  source_contributions,
  "~/CO_CRA/CO_CRA_aggregate/output/acute/sepmain_2021-06-23_source_concentrations.csv",
  row.names = FALSE
)

source_contributions
```




#### graphs for the other separator maintenance scenario (not using these graphs in the MS or supplemental - just using the data; so not updating the graphs) ####



#The separator maintenance trigger canister is the one dated 4/26/2021 from the Livingston sensor (labeled as production)

#plot of total HI vs. distance
```{r}
# define event time
event_time <- as.POSIXct("2021-04-26 10:38", tz = "UTC")

# STEP 1: filter to the event 
plot_df <- hi_final %>%
  filter(`Start Date Time` == event_time)

# optional check: max HI by endpoint
endpoint_max_check <- plot_df %>%
  group_by(Endpoint) %>%
  summarise(
    HI_max = max(HI_total_p95, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(HI_max))

endpoint_max_check

# COLORBLIND-FRIENDLY PALETTE
endpoint_colors <- c(
  "#0072B2", # blue
  "#D55E00", # vermillion
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#CC79A7", # reddish purple
  "#56B4E9", # sky blue
  "#E69F00", # orange
  "#999999"  # gray
)

# repeat colors if there are more endpoints than colors
endpoint_levels <- sort(unique(plot_df$Endpoint))
endpoint_colors <- rep(endpoint_colors, length.out = length(endpoint_levels))
names(endpoint_colors) <- endpoint_levels

# STEP 2: plot all endpoints
hi_distance_all <- ggplot(
  plot_df,
  aes(
    x = distance_ft,
    y = HI_total_p95,
    color = Endpoint,
    group = Endpoint
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(
    x = "Distance from Wellpad (ft)",
    y = "HI (p95 ozone)",
    color = "Endpoint",
    title = "Hazard Index vs Distance"
  ) +
  scale_color_manual(values = endpoint_colors) +
  theme_minimal()

# view plot
hi_distance_all

# STEP 3: save
ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/acute/sepmain_2021-04-26_HIvdistance_all_endpoints.png",
  plot = hi_distance_all,
  width = 8,
  height = 6,
  dpi = 300
)

write.csv(
  plot_df,
  file = "~/CO_CRA/CO_CRA_aggregate/output/acute/sepmain_2021-04-26_HIvdistance_all_endpoints.csv",
  row.names = FALSE
)
```


#plot of chemical contributions
```{r}
# endpoints to plot
endpoints_to_plot <- c(
  "immunological",
  "developmental",
  "hematological",
  "respiratory"
)

# filter HQ dataset for event + selected endpoints
hq_event_df <- hq_all %>%
  filter(
    `Start Date Time` == event_time,
    Endpoint %in% endpoints_to_plot
  ) %>%
  group_by(Operation, Endpoint, distance_ft, chemical) %>%
  summarise(
    HQ_mean = mean(HQ, na.rm = TRUE),
    .groups = "drop"
  )

# clean chemical names
hq_event_df <- hq_event_df %>%
  mutate(
    chemical = recode(
      chemical,
      "aliphatic_low" = "Aliphatic low",
      "aromatic_medium" = "Aromatic medium",
      "benzene" = "Benzene",
      "c2cl4" = "PCE",
      "c2hcl3" = "TCE",
      "ethylbenzene" = "Ethylbenzene",
      "n_nonane" = "n_Nonane",
      "ozone" = "Ozone (p95)",
      "styrene" = "Styrene",
      "toluene" = "Toluene",
      "xylenes" = "Xylenes"
    )
  )

# factor levels
hq_event_df$chemical <- factor(
  hq_event_df$chemical,
  levels = names(chemical_colors)
)

hq_event_df$Endpoint <- factor(
  hq_event_df$Endpoint,
  levels = endpoints_to_plot
)

# plot HQs, faceted by selected endpoints
hq_event_plot <- ggplot(
  hq_event_df,
  aes(
    x = factor(distance_ft),
    y = HQ_mean,
    fill = chemical
  )
) +
  geom_col(position = "stack", width = 0.8) +
  facet_wrap(~ Endpoint) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_fill_manual(
    values = chemical_colors,
    drop = FALSE
  ) +
  labs(
    title = paste("HQ Contributions by Endpoint - Trigger Event", event_time),
    x = "Distance from Wellpad (ft)",
    y = "Hazard Quotient",
    fill = "Chemical"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )

# view and save
hq_event_plot

ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/acute/sepmain_2021-04-26_HQvdistance_selected_endpoints.png",
  plot = hq_event_plot,
  width = 8,
  height = 6,
  dpi = 300
)

# create raw HQ dataset to save for reporting, selected endpoints only
hq_event_raw <- hq_all %>%
  filter(
    `Start Date Time` == event_time,
    Endpoint %in% endpoints_to_plot
  )

write.csv(
  hq_event_raw,
  file = "~/CO_CRA/CO_CRA_aggregate/output/acute/sepmain_2021-04-26_HQs_selected_endpoints.csv",
  row.names = FALSE
)
```




#what fraction of the exposure for this event is attributable to the acute event and the monthly data
```{r}
# Definte event
event_time <- as.POSIXct("2021-04-26 10:38", tz = "UTC")

# Chronic concentrations
chronic_event <- chronic_ready %>%
  filter(`Start Date Time` == event_time) %>%
  group_by(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    chemical
  ) %>%
  summarise(
    chronic_concentration = sum(concentration, na.rm = TRUE),
    .groups = "drop"
  )

# Acute concentrations
acute_event <- acute_clean %>%
  filter(`Start Date Time` == event_time) %>%
  group_by(
    `Start Date Time`,
    `Source Wellpad`,
    `Sampling Site`,
    distance_ft,
    chemical
  ) %>%
  summarise(
    acute_concentration = sum(concentration, na.rm = TRUE),
    .groups = "drop"
  )

# Join together
source_contributions <- full_join(
  acute_event,
  chronic_event,
  by = c(
    "Start Date Time",
    "Source Wellpad",
    "Sampling Site",
    "distance_ft",
    "chemical"
  )
) %>%
  mutate(
    acute_concentration   = coalesce(acute_concentration, 0),
    chronic_concentration = coalesce(chronic_concentration, 0),

    total_concentration =
      acute_concentration + chronic_concentration,

    pct_acute =
      if_else(
        total_concentration > 0,
        100 * acute_concentration / total_concentration,
        NA_real_
      ),

    pct_chronic =
      if_else(
        total_concentration > 0,
        100 * chronic_concentration / total_concentration,
        NA_real_
      )
  ) %>%
  arrange(chemical, distance_ft)

# Save
write.csv(
  source_contributions,
  "~/CO_CRA/CO_CRA_aggregate/output/acute/sepmain_2021-04-26_source_concentrations.csv",
  row.names = FALSE
)

source_contributions
```












