---
title: "chronic_aggregate"
output: html_document
date: "2026-03-10"
---


#load libraries
```{r}
library("dplyr")
library("tidyr")
library("Rmisc")
library("terra")
library("readxl")
library("tidyverse")
library("lubridate")
library("viridis")
detach(package:Rmisc)
detach(package:plyr)
```


#load data
```{r}
rfcs = read.csv("~/CO_CRA/CO_CRA_aggregate/data/RfCs_5-2026.csv")
iurs = read.csv("~/CO_CRA/CO_CRA_aggregate/data/IURs.csv")
dist_source = read.csv("~/CO_CRA/CO_CRA_aggregate/data/broomfield/distance_to_nearest_source.csv")

#ozone
ozone = read.csv("~/CO_CRA/CO_CRA_aggregate/data/Ozone8hour2018to2023.csv")

#load coordinates
x <- as.matrix(read.csv("~/CO_CRA/CO_CRA_aggregate/data//broomfield/XX_coordinates.csv", header = FALSE))
y <- as.matrix(read.csv("~/CO_CRA/CO_CRA_aggregate/data//broomfield/YY_coordinates.csv", header = FALSE))
```


#load montly concentration data
```{r}
data_dir <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield/monthly/gapfilled"

files <- list.files(data_dir, pattern = "\\.xlsx$", full.names = TRUE)

# Extract YYYYMM from filename and sort
file_dates <- stringr::str_extract(files, "\\d{6}")
files <- files[order(file_dates)]

basename(files)

#load all files and sheets
data_list <- list()

for (f in files) {

  sheet_names <- excel_sheets(f)

  chem_list <- list()

  for (s in sheet_names) {
    chem_list[[s]] <- as.matrix(read_excel(f, sheet = s, col_names = FALSE))
  }

  data_list[[basename(f)]] <- chem_list
}

names(data_list) <- file_dates

#checks

# Number of months
length(data_list)

# Chemical names
names(data_list[[1]])
chemicals <- names(data_list[[1]])

# Grid size
dim(data_list[[1]][[1]])

# Check all months have same chemicals
all(sapply(data_list, names) == names(data_list[[1]]))
```


#load annual production concentration data 
```{r}
annual_dir <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield"

annual_files <- list.files(
  annual_dir,
  pattern = "annual_gapfilled_\\d{8}\\.xlsx$",
  full.names = TRUE
)

# extract YYYYMM and sort 
annual_dates <- stringr::str_extract(annual_files, "\\d{6}")
annual_files <- annual_files[order(annual_dates)]

# load all annual files
annual_data_list <- list()

for (f in annual_files) {

  sheet_names <- excel_sheets(f)

  chem_list <- list()

  for (s in sheet_names) {
    chem_list[[s]] <- as.matrix(read_excel(f, sheet = s, col_names = FALSE))
  }

  annual_data_list[[basename(f)]] <- chem_list
}

names(annual_data_list) <- annual_dates

#checks
# number of annual files
length(annual_data_list)

# chemical names match?
all(sapply(annual_data_list, names) == chemicals)

# grid sizes match?
all(sapply(annual_data_list[[1]], dim) == dim(data_list[[1]][[1]]))
```


#average across 4 annual production files for use in TWA
```{r}
# get chemical names
chemicals <- names(data_list[[1]])

# initialize averaged annual list
annual_avg_list <- list()

for (chem in chemicals) {

  # start with zero matrix
  sum_mat <- 0

  for (j in seq_along(annual_data_list)) {
    sum_mat <- sum_mat + annual_data_list[[j]][[chem]]
  }

  # average across years
  annual_avg_list[[chem]] <- sum_mat / length(annual_data_list)
}

#checks
# same dimensions?
all(sapply(annual_avg_list, dim) == dim(data_list[[1]][[1]]))

# check values
summary(as.vector(annual_avg_list[[1]]))
```


#calculate monthly and 30 year ozone - TWA 
```{r}
#format dates
dates <- as.Date(paste0(names(data_list), "01"), format = "%Y%m%d")

days_in_month_vec <- as.numeric(format(
  as.Date(format(dates + 32, "%Y-%m-01")) - 1,
  "%d"
))

#filter
ozone_clean <- ozone %>%
  mutate(
    Date.Local = as.Date(Date.Local),
    year_month = format(Date.Local, "%Y%m"),
    Arithmetic.Mean = Arithmetic.Mean * 1000  # convert ppm → ppb
  )

# keep months that match broomfield data
months_needed <- names(data_list)

ozone_filtered <- ozone_clean %>%
  filter(year_month %in% months_needed)

# monthly median ozone for each month
ozone_monthly <- ozone_filtered %>%
  group_by(year_month) %>%
  summarise(
    monthly_median = median(Arithmetic.Mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year_month)

# check
all(ozone_monthly$year_month == months_needed)

# days per month
ozone_monthly$days <- days_in_month_vec

# monthly contribution 
monthly_num <- sum(ozone_monthly$monthly_median * ozone_monthly$days)
monthly_den <- sum(ozone_monthly$days)

# 30-year contribution 
ozone_longterm_median <- median(ozone_filtered$Arithmetic.Mean, na.rm = TRUE)

annual_days <- 30 * 365

longterm_num <- ozone_longterm_median * annual_days

# ozone TWA (median-based)
ozone_TWA <- (monthly_num + longterm_num) / (monthly_den + annual_days)

# checks
summary(ozone_filtered$Arithmetic.Mean)

ozone_monthly[1:5, ]

ozone_TWA
```


#calculate TWAs for each chemical and grid cell
```{r}
#annual days for production calculation
annual_days <- 30 * 365

#get number of days per month
dates <- as.Date(paste0(names(data_list), "01"), format = "%Y%m%d")

days_in_month <- as.numeric(format(
  as.Date(format(dates + 32, "%Y-%m-01")) - 1,
  "%d"
))

#check
data.frame(
  month = names(data_list)[1:6],
  days = days_in_month[1:6]
)

length(days_in_month) == length(data_list)

#get chemical names
chemicals <- names(data_list[[1]])

## calculate TWA for all chemicals ##
TWA_list <- list()

for (chem in chemicals) {

  # monthly contribution
  num <- 0

  for (i in seq_along(data_list)) {
    num <- num + data_list[[i]][[chem]] * days_in_month[i]
  }

  # add 30-year averaged annual contribution
  num <- num + annual_avg_list[[chem]] * annual_days

  # denominator
  den <- sum(days_in_month) + annual_days

  TWA_list[[chem]] <- num / den
}

#checks
# number of chemicals
length(TWA_list)

# check one
dim(TWA_list[[1]])

# compare to original
dim(data_list[[1]][[1]])
```


#fix chemical names between rfcs and data
```{r}
rfcs_chronic <- rfcs %>%
  dplyr::filter(Exposure == "chronic")

# from TWA data
twa_chems <- names(TWA_list)

# from RfC table
rfc_chems <- unique(rfcs_chronic$Chemical)

twa_chems
rfc_chems

rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Benzene"] <- "benzene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Toluene"] <- "toluene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Ethylbenzene"] <- "ethylbenzene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Xylenes"] <- "xylenes"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "n-nonane"] <- "nonane"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "n-decane"] <- "decane"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Aliphatic low"] <- "aliphatic low"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Aromatic medium"] <- "aromatic medium"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Styrene"] <- "styrene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Trichloroethylene (TCE)"] <- "c2hcl3"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Tetrachloroethylene (PCE)"] <- "c2cl4"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Propene"] <- "propene"

#check
setdiff(names(TWA_list), rfcs_chronic$Chemical)
setdiff(rfcs_chronic$Chemical, names(TWA_list))
```


#calculate spatial chronic HQs for each chemical/endpoint combo in broomfield data
```{r}
endpoints <- unique(rfcs_chronic$Endpoint)
HQ_list <- list()

#loop through and calculate HQ for endpoints and chemicals
for (ep in endpoints) {

  HQ_list[[ep]] <- list()

  for (chem in chemicals) {

    # get RfC for this chemical + endpoint
    rfc_val <- rfcs_chronic %>%
      dplyr::filter(Chemical == chem, Endpoint == ep) %>%
      dplyr::pull(RfC)

    # skip if no RfC exists
    if (length(rfc_val) == 0) next

    # if multiple values exist, take first (or flag later)
    rfc_val <- rfc_val[1]

    # calculate HQ
    HQ_list[[ep]][[chem]] <- TWA_list[[chem]] / rfc_val
  }
}

#checks
HQ_list[["respiratory"]][["benzene"]]

# number of endpoints
length(HQ_list)

# chemicals within an endpoint
names(HQ_list[[1]])

# dimensions match original grid
dim(HQ_list[[1]][[1]])
dim(TWA_list[[1]])

# check values
summary(as.vector(HQ_list[[1]][[1]]))

#missing chemical/rfcs
missing_chems <- setdiff(chemicals, rfcs_chronic$Chemical)
missing_chems

```


#calculate chronic HQs for ozone (not spatial - consistent across area)
```{r}
#create ozone matrix
template_mat <- TWA_list[[1]]

ozone_mat <- matrix(
  ozone_TWA,
  nrow = nrow(template_mat),
  ncol = ncol(template_mat)
)

#calculate ozone HQs and add to HQ_list
for (ep in endpoints) {

  # get ozone RfC for this endpoint
  rfc_val <- rfcs_chronic %>%
    dplyr::filter(Chemical == "Ozone", Endpoint == ep) %>%
    dplyr::pull(RfC)

  # skip if no RfC exists
  if (length(rfc_val) == 0) next

  rfc_val <- rfc_val[1]

  # calculate HQ for ozone
  HQ_ozone <- ozone_mat / rfc_val

  # add to HQ_list
  HQ_list[[ep]][["ozone"]] <- HQ_ozone
}

#check
ep_with_ozone <- names(HQ_list)[
  sapply(HQ_list, function(x) "ozone" %in% names(x))
][1]

HQ_ozone_test <- HQ_list[[ep_with_ozone]][["ozone"]]

#dimensions
dim(HQ_ozone_test)
dim(TWA_list[[1]])
```


#calculate spatial chronic HIs for each endpoint
```{r}
# initialize HI list
HI_list <- list()

for (ep in names(HQ_list)) {

  # get all HQ matrices for this endpoint
  chem_HQs <- HQ_list[[ep]]

  # skip if no chemicals (safety check)
  if (length(chem_HQs) == 0) next

  # sum matrices cell by cell (includes ozone automatically if present)
  HI_list[[ep]] <- Reduce(`+`, chem_HQs)
}

#checks
# number of endpoints
length(HI_list)

# dimensions match original grid
dim(HI_list[[1]])
dim(TWA_list[[1]])

# check values
summary(as.vector(HI_list[[1]]))

#inspection per endpoint
for (ep in names(HI_list)) {
  cat("Endpoint:", ep, "\n")
  print(summary(as.vector(HI_list[[ep]])))
  cat("\n")
}
```


#remove any grids within 25 m or less than the sampling source
```{r}
# convert distance dataframe to matrix
dist_mat <- as.matrix(dist_source)

mask <- dist_mat >= 25   # TRUE = keep, FALSE = remove

#apply mask to HI results
HI_list_masked <- list()

for (ep in names(HI_list)) {
  HI_list[[ep]] <- HI_list[[ep]]
  HI_list[[ep]][!mask] <- NA
}

#checks
# how many cells removed?
sum(!mask)

# visualize mask
image(mask)
```


#summary stats for chronic HIs
```{r}
# function to calculate summary stats
calc_summary <- function(mat) {

  vals <- as.vector(mat)

  vals <- vals[!is.na(vals)]

  data.frame(
    min = min(vals),
    mean = mean(vals),
    median = median(vals),
    max = max(vals),
    sd = sd(vals),
    se = sd(vals) / sqrt(length(vals))
  )
}

# apply to all endpoints
HI_summary <- do.call(
  rbind,
  lapply(names(HI_list), function(ep) {
    cbind(
      endpoint = ep,
      calc_summary(HI_list[[ep]])
    )
  })
)

# clean row names
rownames(HI_summary) <- NULL

HI_summary

write.csv(HI_summary, file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/HI_summary.csv")
```


#spatial plots of HI per endpoint
```{r}
# function to convert matrix + coords to df
make_df <- function(mat, x, y) {

  data.frame(
    x = as.vector(x),
    y = as.vector(y),
    HI = as.vector(mat)
  )
}

#define bounds
x_min <- 496000
x_max <- 501000

y_min <- 4425000
y_max <- 4426500

# load well pad coordinates
wellpad_cords <- read.csv(
  "~/CO_CRA/CO_CRA_aggregate/data/broomfield/wellpad_cords.csv"
)

# create 610 m radius circles around each well pad
radius <- 610

circle_df <- do.call(
  rbind,
  lapply(seq_len(nrow(wellpad_cords)), function(i) {

    theta <- seq(0, 2 * pi, length.out = 200)

    data.frame(
      id = i,
      x = wellpad_cords$X[i] + radius * cos(theta),
      y = wellpad_cords$Y[i] + radius * sin(theta)
    )
  })
)

#define legend
hi_min <- 0
hi_max <- 0.71

hi_breaks <- seq(hi_min, hi_max, by = 0.1)

# plot all endpoints
plots <- lapply(names(HI_list), function(ep) {

  df <- make_df(HI_list[[ep]], x, y)

  ggplot(df, aes(x = x, y = y, fill = HI)) +

    geom_raster() +

    # 610 m setback circles
    geom_path(
      data = circle_df,
      aes(x = x, y = y, group = id),
      color = "black",
      linewidth = 0.8,
      inherit.aes = FALSE
    ) +

    scale_fill_viridis_c(
      name = "HI",
      limits = c(hi_min, hi_max),
      breaks = hi_breaks,
      oob = scales::squish
    ) +

    coord_cartesian(
      xlim = c(x_min, x_max),
      ylim = c(y_min, y_max)
    ) +

    labs(
      title = paste("Hazard Index -", ep),
      x = "X (m)",
      y = "Y (m)"
    ) +

    theme_minimal()

})

# view plots
plots[[1]]
plots[[2]]
plots[[3]]
plots[[4]]
plots[[5]]
plots[[6]]
plots[[7]]
plots[[8]]
plots[[9]]
plots[[10]]
plots[[11]]

# save plots
for (i in seq_along(plots)) {

  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/chronic",
      paste0("HI_", names(HI_list)[i], ".png")
    ),
    plot = plots[[i]],
    width = 6,
    height = 5,
    dpi = 300
  )
}
```



### cancer analysis ###




#fix chemical names between iurs and data
```{r}
# check chemical names
unique(iurs$Chemical)
names(TWA_list)

# fix mismatches 
iurs$Chemical[iurs$Chemical == "Benzene"] <- "benzene"
iurs$Chemical[iurs$Chemical == "Ethylbenzene"] <- "ethylbenzene"
iurs$Chemical[iurs$Chemical == "C2HCl3"] <- "c2hcl3"
iurs$Chemical[iurs$Chemical == "C2Cl4"] <- "c2cl4"
```


#calculate excess cancer risk, assuming 33 years of exposure (3 years of data in pre-production and 30 in production)
```{r}
CancerRisk_list <- list()

# exposure duration adjustment
ED_adjust <- 33 / 78

#formula
for (chem in names(TWA_list)) {
  
  # get IUR for this chemical
  iur_val <- iurs %>%
    dplyr::filter(Chemical == chem) %>%
    dplyr::pull(IUR..ppb.)
  
  # skip if no IUR
  if (length(iur_val) == 0) next
  
  iur_val <- iur_val[1]
  
  # calculate risk with exposure duration adjustment
  CancerRisk_list[[chem]] <- TWA_list[[chem]] * iur_val * ED_adjust
}

#sum cancer risk across chemicals
CancerRisk_total <- Reduce(`+`, CancerRisk_list)

#apply the <25 m mask
CancerRisk_total[!mask] <- NA

#checks
# dimensions
dim(CancerRisk_total)
dim(TWA_list[[1]])

# summary
summary(as.vector(CancerRisk_total))

# inspect one chemical contribution
summary(as.vector(CancerRisk_list[[1]]))

# save total cancer risk matrix for use in calculating background-only cancer risk in _wbkg Rmd
saveRDS(
  CancerRisk_total,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/CancerRisk_total_nobkg.rds"
)

# save individual chemical cancer risks for use in calculating background-only cancer risk in _wbkg Rmd
saveRDS(
  CancerRisk_list,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/CancerRisk_list_nobkg.rds"
)
```


#summary stats
```{r}
#total cancer risk
cancer_summary_tot <- calc_summary(CancerRisk_total)

cancer_summary_tot

write.csv(
  cancer_summary_tot,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_tot.csv",
  row.names = FALSE
)

#summary per chemical
cancer_summary_ind <- do.call(
  rbind,
  lapply(names(CancerRisk_list), function(chem) {
    cbind(
      chemical = chem,
      calc_summary(CancerRisk_list[[chem]])
    )
  })
)

# clean row names
rownames(cancer_summary_ind) <- NULL

cancer_summary_ind

write.csv(
  cancer_summary_ind,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_ind.csv",
  row.names = FALSE
)
```


#map cancer risks
```{r}
make_df <- function(mat, x, y) {
  data.frame(
    x = as.vector(x),
    y = as.vector(y),
    Risk = as.vector(mat)
  )
}

# fixed/defined legend
risk_min <- 8.5e-8
risk_max <- 5.2e-4

risk_breaks <- c(1e-7, 1e-6, 1e-5, 1e-4, 5e-4)

# regulatory limits
contour_levels <- c(1e-6, 1e-4)

# create data
df <- make_df(CancerRisk_total, x, y)

# plot
plot_cancer_tot <- ggplot(df, aes(x = x, y = y, fill = Risk)) +
  
  geom_raster() +
  
  scale_fill_viridis_c(
    name = "Cancer Risk",
    trans = "log10",
    limits = c(risk_min, risk_max),
    breaks = risk_breaks,
    oob = scales::squish
  ) +
  
  coord_cartesian(
    xlim = c(x_min, x_max),
    ylim = c(y_min, y_max)
  ) +
  
  geom_contour(
    aes(z = Risk),
    breaks = contour_levels,
    color = "red",
    linewidth = 0.7
  ) +
  
  # 610 m setback circles
  geom_path(
    data = circle_df,
    aes(x = x, y = y, group = id),
    color = "black",
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  
  labs(
    title = "Total Excess Cancer Risk",
    x = "X (m)",
    y = "Y (m)"
  ) +
  
  theme_minimal()

# view
plot_cancer_tot

# save
ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_tot.png",
  plot = plot_cancer_tot,
  width = 6,
  height = 5,
  dpi = 300
)
```

#map of just benzene to compare (not using anymore, so not editing legend, etc.)
```{r}
# regulatory limits
contour_levels <- c(1e-6, 1e-5, 1e-4)

# --- select benzene ---
benzene_mat <- CancerRisk_list[["benzene"]]

# create dataframe
df <- make_df(benzene_mat, x, y)

plot_cancer_benzene <- ggplot(df, aes(x = x, y = y, fill = Risk)) +
  geom_raster() +
  scale_fill_viridis_c(
    name = "Cancer Risk",
    trans = "log10"
  ) +
  coord_cartesian(
    xlim = c(x_min, x_max),
    ylim = c(y_min, y_max)
  ) +
  geom_contour(
    aes(z = Risk),
    breaks = contour_levels,
    color = "red",
    linewidth = 0.7
  ) +
  labs(
    title = "Excess Cancer Risk - Benzene",
    x = "X (m)",
    y = "Y (m)"
  ) +
  theme_minimal()

plot_cancer_benzene

ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_benzene.png",
  plot = plot_cancer_benzene,
  width = 6,
  height = 5,
  dpi = 300
)
```























