---
title: "chronic aggregate analysis with background"
output: html_document
date: "2026-04-01"
---


#load libraries
```{r}
library("dplyr")
library("tidyr")
library("Rmisc")
library("terra")
library("readxl")
library("tidyverse")
library("lubridate")
library("viridis")
library("sf")
library("openxlsx")
library("patchwork")
library("ggplot2")
detach(package:Rmisc)
detach(package:plyr)
```


#load data
```{r}
rfcs = read.csv("~/CO_CRA/CO_CRA_aggregate/data/RfCs_5-2026.csv")
iurs = read.csv("~/CO_CRA/CO_CRA_aggregate/data/IURs.csv")
dist_source = read.csv("~/CO_CRA/CO_CRA_aggregate/data/broomfield/distance_to_nearest_source.csv")

#ozone
ozone = read.csv("~/CO_CRA/CO_CRA_aggregate/data/Ozone8hour2018to2023.csv")

#load coordinates
x <- as.matrix(read.csv("~/CO_CRA/CO_CRA_aggregate/data//broomfield/XX_coordinates.csv", header = FALSE))
y <- as.matrix(read.csv("~/CO_CRA/CO_CRA_aggregate/data//broomfield/YY_coordinates.csv", header = FALSE))
```



#load montly concentration data
```{r}
data_dir <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield/monthly/gapfilled_wbkg"

files <- list.files(data_dir, pattern = "\\.xlsx$", full.names = TRUE)

# Extract YYYYMM from filename and sort
file_dates <- stringr::str_extract(files, "\\d{6}")
files <- files[order(file_dates)]

basename(files)

#load all files and sheets
data_list <- list()

for (f in files) {

  sheet_names <- excel_sheets(f)

  chem_list <- list()

  for (s in sheet_names) {
    chem_list[[s]] <- as.matrix(read_excel(f, sheet = s, col_names = FALSE))
  }

  data_list[[basename(f)]] <- chem_list
}

names(data_list) <- file_dates

#checks

# Number of months
length(data_list)

# Chemical names
names(data_list[[1]])
chemicals <- names(data_list[[1]])

# Grid size
dim(data_list[[1]][[1]])

# Check all months have same chemicals
all(sapply(data_list, names) == names(data_list[[1]]))
```


#load annual production concentration data
```{r}
annual_dir <- "~/CO_CRA/CO_CRA_aggregate/data/broomfield"

annual_files <- list.files(
  annual_dir,
  pattern = "annual_gapfilled_with_bkg_\\d{8}\\.xlsx$",
  full.names = TRUE
)

# extract YYYYMM and sort 
annual_dates <- stringr::str_extract(annual_files, "\\d{6}")
annual_files <- annual_files[order(annual_dates)]

# load all annual files
annual_data_list <- list()

for (f in annual_files) {

  sheet_names <- excel_sheets(f)

  chem_list <- list()

  for (s in sheet_names) {
    chem_list[[s]] <- as.matrix(read_excel(f, sheet = s, col_names = FALSE))
  }

  annual_data_list[[basename(f)]] <- chem_list
}

names(annual_data_list) <- annual_dates

#checks
# number of annual files
length(annual_data_list)

# chemical names match?
all(sapply(annual_data_list, names) == chemicals)

# grid sizes match?
all(sapply(annual_data_list[[1]], dim) == dim(data_list[[1]][[1]]))
```


#average across 4 annual production files for use in TWA
```{r}
# get chemical names
chemicals <- names(data_list[[1]])

# initialize averaged annual list
annual_avg_list <- list()

for (chem in chemicals) {

  # start with zero matrix
  sum_mat <- 0

  for (j in seq_along(annual_data_list)) {
    sum_mat <- sum_mat + annual_data_list[[j]][[chem]]
  }

  # average across years
  annual_avg_list[[chem]] <- sum_mat / length(annual_data_list)
}

#checks
# same dimensions?
all(sapply(annual_avg_list, dim) == dim(data_list[[1]][[1]]))

# check values
summary(as.vector(annual_avg_list[[1]]))

## save data for reporting ##

# create workbook
wb <- createWorkbook()

for (chem in names(annual_avg_list)) {

  # add worksheet
  addWorksheet(wb, sheetName = substr(chem, 1, 31))

  # convert matrix to dataframe
  chem_df <- as.data.frame(annual_avg_list[[chem]])

  # write to worksheet
  writeData(wb, sheet = substr(chem, 1, 31), x = chem_df)
}

# save workbook
saveWorkbook(
  wb,
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/annual_average_chemical_grids_wbkg.xlsx",
  overwrite = TRUE
)

```



#calculate monthly and 30 year ozone - TWA 
```{r}
#format dates
dates <- as.Date(paste0(names(data_list), "01"), format = "%Y%m%d")

days_in_month_vec <- as.numeric(format(
  as.Date(format(dates + 32, "%Y-%m-01")) - 1,
  "%d"
))

#filter
ozone_clean <- ozone %>%
  mutate(
    Date.Local = as.Date(Date.Local),
    year_month = format(Date.Local, "%Y%m"),
    Arithmetic.Mean = Arithmetic.Mean * 1000  # convert ppm → ppb
  )

# keep months that match broomfield data
months_needed <- names(data_list)

ozone_filtered <- ozone_clean %>%
  filter(year_month %in% months_needed)

# monthly median ozone for each month
ozone_monthly <- ozone_filtered %>%
  group_by(year_month) %>%
  summarise(
    monthly_median = median(Arithmetic.Mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(year_month)

# check
all(ozone_monthly$year_month == months_needed)

# days per month
ozone_monthly$days <- days_in_month_vec

# monthly contribution 
monthly_num <- sum(ozone_monthly$monthly_median * ozone_monthly$days)
monthly_den <- sum(ozone_monthly$days)

# 30-year contribution 
ozone_longterm_median <- median(ozone_filtered$Arithmetic.Mean, na.rm = TRUE)

annual_days <- 30 * 365

longterm_num <- ozone_longterm_median * annual_days

# ozone TWA (median-based)
ozone_TWA <- (monthly_num + longterm_num) / (monthly_den + annual_days)

# checks
summary(ozone_filtered$Arithmetic.Mean)

ozone_monthly[1:5, ]

ozone_TWA
```




#calculate TWAs for each chemical and grid cell
```{r}
#annual days for production calculation
annual_days <- 30 * 365

#get number of days per month
dates <- as.Date(paste0(names(data_list), "01"), format = "%Y%m%d")

days_in_month <- as.numeric(format(
  as.Date(format(dates + 32, "%Y-%m-01")) - 1,
  "%d"
))

#check
data.frame(
  month = names(data_list)[1:6],
  days = days_in_month[1:6]
)

length(days_in_month) == length(data_list)

#get chemical names
chemicals <- names(data_list[[1]])

## calculate TWA for all chemicals ##
TWA_list <- list()

for (chem in chemicals) {

  # monthly contribution
  num <- 0

  for (i in seq_along(data_list)) {
    num <- num + data_list[[i]][[chem]] * days_in_month[i]
  }

  # add 30-year averaged annual contribution
  num <- num + annual_avg_list[[chem]] * annual_days

  # denominator
  den <- sum(days_in_month) + annual_days

  TWA_list[[chem]] <- num / den
}

#checks
# number of chemicals
length(TWA_list)

# check one
dim(TWA_list[[1]])

# compare to original
dim(data_list[[1]][[1]])

## save data for reporting ##

wb <- createWorkbook()

for (chem in names(TWA_list)) {

  addWorksheet(wb, sheetName = substr(chem, 1, 31))

  chem_df <- as.data.frame(TWA_list[[chem]])

  writeData(
    wb,
    sheet = substr(chem, 1, 31),
    x = chem_df
  )
}

saveWorkbook(
  wb,
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/TWA_chemical_grids_wbkg.xlsx",
  overwrite = TRUE
)
```



#fix chemical names between rfcs and data
```{r}
rfcs_chronic <- rfcs %>%
  dplyr::filter(Exposure == "chronic")

# from TWA data
twa_chems <- names(TWA_list)

# from RfC table
rfc_chems <- unique(rfcs_chronic$Chemical)

twa_chems
rfc_chems

rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Benzene"] <- "benzene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Toluene"] <- "toluene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Ethylbenzene"] <- "ethylbenzene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Xylenes"] <- "xylenes"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "n-nonane"] <- "nonane"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "n-decane"] <- "decane"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Aliphatic low"] <- "aliphatic low"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Aromatic medium"] <- "aromatic medium"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Styrene"] <- "styrene"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Trichloroethylene (TCE)"] <- "c2hcl3"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Tetrachloroethylene (PCE)"] <- "c2cl4"
rfcs_chronic$Chemical[rfcs_chronic$Chemical == "Propene"] <- "propene"

#check
setdiff(names(TWA_list), rfcs_chronic$Chemical)
setdiff(rfcs_chronic$Chemical, names(TWA_list))
```



#calculate spatial chronic HQs for each chemical/endpoint combo in broomfield data
```{r}
endpoints <- unique(rfcs_chronic$Endpoint)
HQ_list <- list()

#loop through and calculate HQ for endpoints and chemicals
for (ep in endpoints) {

  HQ_list[[ep]] <- list()

  for (chem in chemicals) {

    # get RfC for this chemical + endpoint
    rfc_val <- rfcs_chronic %>%
      dplyr::filter(Chemical == chem, Endpoint == ep) %>%
      dplyr::pull(RfC)

    # skip if no RfC exists
    if (length(rfc_val) == 0) next

    # if multiple values exist, take first (or flag later)
    rfc_val <- rfc_val[1]

    # calculate HQ
    HQ_list[[ep]][[chem]] <- TWA_list[[chem]] / rfc_val
  }
}

#checks
HQ_list[["respiratory"]][["benzene"]]

# number of endpoints
length(HQ_list)

# chemicals within an endpoint
names(HQ_list[[1]])

# dimensions match original grid
dim(HQ_list[[1]][[1]])
dim(TWA_list[[1]])

# check values
summary(as.vector(HQ_list[[1]][[1]]))

#missing chemical/rfcs
missing_chems <- setdiff(chemicals, rfcs_chronic$Chemical)
missing_chems

```


#calculate chronic HQs for ozone (not spatial - consistent across area)
```{r}
#create ozone matrix
template_mat <- TWA_list[[1]]

ozone_mat <- matrix(
  ozone_TWA,
  nrow = nrow(template_mat),
  ncol = ncol(template_mat)
)

#calculate ozone HQs and add to HQ_list
for (ep in endpoints) {

  # get ozone RfC for this endpoint
  rfc_val <- rfcs_chronic %>%
    dplyr::filter(Chemical == "Ozone", Endpoint == ep) %>%
    dplyr::pull(RfC)

  # skip if no RfC exists
  if (length(rfc_val) == 0) next

  rfc_val <- rfc_val[1]

  # calculate HQ for ozone
  HQ_ozone <- ozone_mat / rfc_val

  # add to HQ_list
  HQ_list[[ep]][["ozone"]] <- HQ_ozone
}

#check
ep_with_ozone <- names(HQ_list)[
  sapply(HQ_list, function(x) "ozone" %in% names(x))
][1]

HQ_ozone_test <- HQ_list[[ep_with_ozone]][["ozone"]]

#dimensions
dim(HQ_ozone_test)
dim(TWA_list[[1]])
```

#save HQ files for reporting, saving an HQ file for each endpoint with chemicals in different tabs in the excel file, including ozone
```{r}
output_dir <- "~/CO_CRA/CO_CRA_aggregate/output/chronic/raw_HQs"

for (ep in endpoints) {

  wb <- createWorkbook()

  for (chem in names(HQ_list[[ep]])) {

    if (is.null(HQ_list[[ep]][[chem]])) next

    sheet_name <- substr(chem, 1, 31)

    addWorksheet(wb, sheet_name)

    chem_df <- as.data.frame(HQ_list[[ep]][[chem]])

    writeData(
      wb,
      sheet = sheet_name,
      x = chem_df
    )
  }

  file_name <- paste0("HQ_chronic_wbkg_", ep, ".xlsx")

  saveWorkbook(
    wb,
    file.path(output_dir, file_name),
    overwrite = TRUE
  )
}
```



#calculate spatial chronic HIs for each endpoint
```{r}
# initialize HI list
HI_list <- list()

for (ep in names(HQ_list)) {

  # get all HQ matrices for this endpoint
  chem_HQs <- HQ_list[[ep]]

  # skip if no chemicals (safety check)
  if (length(chem_HQs) == 0) next

  # sum matrices cell by cell (includes ozone automatically if present)
  HI_list[[ep]] <- Reduce(`+`, chem_HQs)
}

#checks
# number of endpoints
length(HI_list)

# dimensions match original grid
dim(HI_list[[1]])
dim(TWA_list[[1]])

# check values
summary(as.vector(HI_list[[1]]))

#inspection per endpoint
for (ep in names(HI_list)) {
  cat("Endpoint:", ep, "\n")
  print(summary(as.vector(HI_list[[ep]])))
  cat("\n")
}

## save data for reporting in one file with endpoint separated by tabs ##

output_file <- "~/CO_CRA/CO_CRA_aggregate/output/chronic/chronic_HIs_wbkg.xlsx"

wb <- createWorkbook()

for (ep in names(HI_list)) {

  # skip empty endpoints (safety)
  if (is.null(HI_list[[ep]])) next

  # create sheet (truncate to Excel limit)
  sheet_name <- substr(ep, 1, 31)

  addWorksheet(wb, sheet_name)

  # convert matrix → data.frame for Excel
  hi_df <- as.data.frame(HI_list[[ep]])

  writeData(
    wb,
    sheet = sheet_name,
    x = hi_df
  )
}

saveWorkbook(
  wb,
  file = output_file,
  overwrite = TRUE
)
```


#remove any grids within 25 m or less than the sampling source
```{r}
# convert distance dataframe to matrix
dist_mat <- as.matrix(dist_source)

mask <- dist_mat >= 25   # TRUE = keep, FALSE = remove

#apply mask to HI results
HI_list_masked <- list()

for (ep in names(HI_list)) {
  HI_list[[ep]] <- HI_list[[ep]]
  HI_list[[ep]][!mask] <- NA
}

#checks
# how many cells removed?
sum(!mask)

# visualize mask
image(mask)
```


#summary stats for chronic HIs
```{r}
# function to calculate summary stats
calc_summary <- function(mat) {

  vals <- as.vector(mat)

  vals <- vals[!is.na(vals)]

  data.frame(
    min = min(vals),
    mean = mean(vals),
    median = median(vals),
    max = max(vals),
    sd = sd(vals),
    se = sd(vals) / sqrt(length(vals))
  )
}

# apply to all endpoints
HI_summary <- do.call(
  rbind,
  lapply(names(HI_list), function(ep) {
    cbind(
      endpoint = ep,
      calc_summary(HI_list[[ep]])
    )
  })
)

# clean row names
rownames(HI_summary) <- NULL

HI_summary

write.csv(HI_summary, file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/HI_summary_wbkg.csv")
```


#spatial plots of HI per endpoint
```{r}
# function to convert matrix + coords to df
make_df <- function(mat, x, y) {

  data.frame(
    x = as.vector(x),
    y = as.vector(y),
    HI = as.vector(mat)
  )
}

#define bounds
x_min <- 496000
x_max <- 501000

y_min <- 4425000
y_max <- 4426500

# read well pad coordinates
wellpad_cords <- read.csv("~/CO_CRA/CO_CRA_aggregate/data/broomfield/wellpad_cords.csv")

# create 610 m radius circles around each well pad
radius <- 610

circle_df <- do.call(
  rbind,
  lapply(seq_len(nrow(wellpad_cords)), function(i) {

    theta <- seq(0, 2 * pi, length.out = 200)

    data.frame(
      id = i,
      x = wellpad_cords$X[i] + radius * cos(theta),
      y = wellpad_cords$Y[i] + radius * sin(theta)
    )
  })
)

#define legend
hi_min <- 0
hi_max <- 0.71

hi_breaks <- seq(hi_min, hi_max, by = 0.1)

# plot all endpoints
plots <- lapply(names(HI_list), function(ep) {

  df <- make_df(HI_list[[ep]], x, y)

  ggplot(df, aes(x = x, y = y, fill = HI)) +

    geom_raster() +

    # 610 m setback circles
    geom_path(
      data = circle_df,
      aes(x = x, y = y, group = id),
      color = "black",
      linewidth = 0.8,
      inherit.aes = FALSE
    ) +

    scale_fill_viridis_c(
      name = "HI",
      limits = c(hi_min, hi_max),
      breaks = hi_breaks,
      oob = scales::squish
    ) +

    coord_cartesian(
      xlim = c(x_min, x_max),
      ylim = c(y_min, y_max)
    ) +

    labs(
      title = paste("Hazard Index -", ep),
      x = "X (m)",
      y = "Y (m)"
    ) +

    theme_minimal()

})

# view plots
plots[[1]]
plots[[2]]
plots[[3]]
plots[[4]]
plots[[5]]
plots[[6]]
plots[[7]]
plots[[8]]
plots[[9]]
plots[[10]]
plots[[11]]

# save plots
for (i in seq_along(plots)) {

  ggsave(
    filename = file.path(
      "~/CO_CRA/CO_CRA_aggregate/output/chronic",
      paste0("HI_wbkg_", names(HI_list)[i], ".png")
    ),
    plot = plots[[i]],
    width = 6,
    height = 5,
    dpi = 300
  )
}
```


### getting range of HIs per endpoint at 2,000 ft ###


#filter for only grids that overlap with the 2,000 ft setback distance 
```{r}
# target radius
radius <- 610

# grid resolution
grid_res <- 50
half_res <- grid_res / 2

# make grid-cell polygons from grid-center coordinates
grid_cells <- data.frame(
  cell_id = seq_along(as.vector(x)),
  x = as.vector(x),
  y = as.vector(y)
) %>%
  rowwise() %>%
  mutate(
    geometry = list(
      st_polygon(list(matrix(
        c(
          x - half_res, y - half_res,
          x + half_res, y - half_res,
          x + half_res, y + half_res,
          x - half_res, y + half_res,
          x - half_res, y - half_res
        ),
        ncol = 2,
        byrow = TRUE
      )))
    )
  ) %>%
  ungroup() %>%
  st_as_sf(crs = NA)

# make wellpad points
wellpad_sf <- st_as_sf(
  wellpad_cords,
  coords = c("X", "Y"),
  crs = NA
)

# create 610 m circle boundaries only, not filled buffers
circle_lines <- wellpad_sf %>%
  st_buffer(dist = radius) %>%
  st_boundary()

# identify grid cells that intersect any 610 m circle line
intersects_610m <- st_intersects(grid_cells, circle_lines)

grid_cells$at_610m <- lengths(intersects_610m) > 0

# check number of selected cells
sum(grid_cells$at_610m)
```


#get the range of HIs for each endpoint in these grids for reporting
```{r}
calc_HI_at_610m <- function(mat, ep) {

  vals <- as.vector(mat)

  vals_610m <- vals[grid_cells$at_610m]

  vals_610m <- vals_610m[!is.na(vals_610m)]

  data.frame(
    endpoint = ep,
    n_grids = length(vals_610m),
    min_HI = min(vals_610m),
    median_HI = median(vals_610m),
    mean_HI = mean(vals_610m),
    max_HI = max(vals_610m)
  )
}

HI_610m_summary <- do.call(
  rbind,
  lapply(names(HI_list), function(ep) {
    calc_HI_at_610m(HI_list[[ep]], ep)
  })
)

rownames(HI_610m_summary) <- NULL

HI_610m_summary

write.csv(
  HI_610m_summary,
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/HI_summary_at_2000ft_wbkg.csv",
  row.names = FALSE
)
```




#summary of background-only HIs at 2,000 ft 
```{r}
# files
# files
hi_wbkg_file <- "~/CO_CRA/CO_CRA_aggregate/output/chronic/chronic_HIs_wbkg.xlsx"
hi_nobkg_file <- "~/CO_CRA/CO_CRA_aggregate/output/chronic/chronic_HIs.xlsx"

# endpoints/sheets
endpoints <- intersect(
  excel_sheets(hi_wbkg_file),
  excel_sheets(hi_nobkg_file)
)

# summary function
calc_summary_at_610m <- function(mat, at_610m, endpoint_name) {

  vals <- as.vector(mat)
  vals <- vals[at_610m]
  vals <- vals[!is.na(vals)]

  data.frame(
    endpoint = endpoint_name,
    n_grids = length(vals),
    min_HI = min(vals),
    median_HI = median(vals),
    mean_HI = mean(vals),
    max_HI = max(vals)
  )
}

# read HI matrix function
read_hi_matrix <- function(file, sheet) {

  df <- read_excel(
    file,
    sheet = sheet,
    col_names = FALSE
  )

  df <- df %>%
    mutate(across(everything(), as.numeric))

  as.matrix(df)
}

# calculate background-only summary one endpoint at a time
HI_background_610m_summary <- do.call(
  rbind,
  lapply(endpoints, function(ep) {

    hi_wbkg <- read_hi_matrix(hi_wbkg_file, ep)
    hi_nobkg <- read_hi_matrix(hi_nobkg_file, ep)

    stopifnot(dim(hi_wbkg) == dim(hi_nobkg))

    hi_bkg <- hi_wbkg - hi_nobkg

    calc_summary_at_610m(
      mat = hi_bkg,
      at_610m = grid_cells$at_610m,
      endpoint_name = ep
    )
  })
)

HI_background_610m_summary

write.csv(
  HI_background_610m_summary,
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/HI_summary_at_2000ft_bkgonly.csv",
  row.names = FALSE
)
```





### chronic HI summary table for all endpoints at all distances, for UOGD, background, total ###



#create spatial boundaries and extract grids that match them
```{r}

# ------------------------------------------------------------
# 1. Define input and output files
# ------------------------------------------------------------

chronic_hi_wbkg_file <-
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/chronic_HIs_wbkg.xlsx"

chronic_hi_nobkg_file <-
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/chronic_HIs.xlsx"

chronic_hi_distance_output_csv <-
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/HI_summary_500_to_5000ft_all_scenarios.csv"

chronic_hi_distance_output_xlsx <-
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/HI_summary_500_to_5000ft_all_scenarios.xlsx"


# ------------------------------------------------------------
# 2. Define distances
# ------------------------------------------------------------

distance_lookup <- data.frame(
  distance_ft = seq(
    from = 500,
    to = 5000,
    by = 500
  )
) %>%
  mutate(
    distance_m = distance_ft * 0.3048
  )

distance_lookup

# Expected distances:
# 500, 1000, 1500, 2000, 2500,
# 3000, 3500, 4000, 4500, 5000 ft


# ------------------------------------------------------------
# 3. Load spatial information
#
# These are loaded into newly named objects so the existing
# x, y, dist_source, grid_cells, and wellpad_sf objects are
# not modified.
# ------------------------------------------------------------

chronic_x <- as.matrix(
  read.csv(
    "~/CO_CRA/CO_CRA_aggregate/data/broomfield/XX_coordinates.csv",
    header = FALSE
  )
)

chronic_y <- as.matrix(
  read.csv(
    "~/CO_CRA/CO_CRA_aggregate/data/broomfield/YY_coordinates.csv",
    header = FALSE
  )
)

chronic_dist_source <- as.matrix(
  read.csv(
    "~/CO_CRA/CO_CRA_aggregate/data/broomfield/distance_to_nearest_source.csv"
  )
)

chronic_wellpad_coords <- read.csv(
  "~/CO_CRA/CO_CRA_aggregate/data/broomfield/wellpad_cords.csv"
)


# ------------------------------------------------------------
# 4. Check that the spatial matrices have matching dimensions
# ------------------------------------------------------------

stopifnot(
  identical(
    dim(chronic_x),
    dim(chronic_y)
  )
)

stopifnot(
  identical(
    dim(chronic_x),
    dim(chronic_dist_source)
  )
)

chronic_grid_nrow <- nrow(chronic_x)
chronic_grid_ncol <- ncol(chronic_x)
chronic_grid_ncells <- length(chronic_x)

data.frame(
  n_rows = chronic_grid_nrow,
  n_columns = chronic_grid_ncol,
  n_cells = chronic_grid_ncells
)


# ------------------------------------------------------------
# 5. Create the <25 m exclusion mask
#
# This recreates the same exclusion used in the main workflow.
# It is applied to all three scenarios after reading the files.
# ------------------------------------------------------------

chronic_keep_mask <- chronic_dist_source >= 25

sum(chronic_keep_mask, na.rm = TRUE)
sum(!chronic_keep_mask, na.rm = TRUE)


# ------------------------------------------------------------
# 6. Build grid-cell polygons
#
# The concentration/HIs represent 50 m x 50 m grid cells.
# The x and y values are treated as grid-cell centers.
# ------------------------------------------------------------

chronic_grid_res <- 50
chronic_half_res <- chronic_grid_res / 2

chronic_grid_df <- data.frame(
  cell_id = seq_len(chronic_grid_ncells),
  x = as.vector(chronic_x),
  y = as.vector(chronic_y)
)

chronic_grid_geometry <- lapply(
  seq_len(nrow(chronic_grid_df)),
  function(i) {

    cell_x <- chronic_grid_df$x[i]
    cell_y <- chronic_grid_df$y[i]

    st_polygon(
      list(
        matrix(
          c(
            cell_x - chronic_half_res,
            cell_y - chronic_half_res,

            cell_x + chronic_half_res,
            cell_y - chronic_half_res,

            cell_x + chronic_half_res,
            cell_y + chronic_half_res,

            cell_x - chronic_half_res,
            cell_y + chronic_half_res,

            cell_x - chronic_half_res,
            cell_y - chronic_half_res
          ),
          ncol = 2,
          byrow = TRUE
        )
      )
    )
  }
)

chronic_grid_cells_sf <- st_sf(
  chronic_grid_df,
  geometry = st_sfc(chronic_grid_geometry)
)


# ------------------------------------------------------------
# 7. Create well-pad spatial points
# ------------------------------------------------------------

chronic_wellpad_sf <- st_as_sf(
  chronic_wellpad_coords,
  coords = c("X", "Y"),
  crs = NA
)


# ------------------------------------------------------------
# 8. Create one grid-cell selection mask per distance
#
# A cell is selected when its polygon intersects the circle
# boundary at the specified distance around any well pad.
# ------------------------------------------------------------

chronic_distance_masks <- setNames(
  vector(
    mode = "list",
    length = nrow(distance_lookup)
  ),
  distance_lookup$distance_ft
)

for (i in seq_len(nrow(distance_lookup))) {

  current_distance_ft <- distance_lookup$distance_ft[i]
  current_distance_m <- distance_lookup$distance_m[i]

  # Create separate circle boundaries around all well pads
  current_circle_lines <- chronic_wellpad_sf %>%
    st_buffer(
      dist = current_distance_m
    ) %>%
    st_boundary()

  # Determine which grid polygons intersect at least one line
  current_intersections <- st_intersects(
    chronic_grid_cells_sf,
    current_circle_lines
  )

  current_mask <- lengths(current_intersections) > 0

  chronic_distance_masks[[
    as.character(current_distance_ft)
  ]] <- current_mask
}


# ------------------------------------------------------------
# 9. Check how many unique grid cells are selected at each
#    distance before applying missing-HI exclusions
# ------------------------------------------------------------

distance_grid_counts <- data.frame(
  distance_ft = distance_lookup$distance_ft,
  distance_m = distance_lookup$distance_m,
  n_spatial_grids = vapply(
    chronic_distance_masks,
    sum,
    numeric(1)
  )
)

distance_grid_counts
```



#formatting and creating functions to get the summary data for the 3 scenarios: 1) UOGD, 2) background, 3) total
```{r}

# ------------------------------------------------------------
# 10. Function to safely read an HI matrix from Excel
#
# The workbooks were created using openxlsx::writeData(),
# which writes column names by default. Therefore:
#
#   colNames = TRUE
#
# is used here.
#
# skipEmptyRows and skipEmptyCols are disabled to preserve
# the original matrix dimensions.
# ------------------------------------------------------------

read_chronic_hi_matrix <- function(
  file,
  sheet,
  expected_nrow,
  expected_ncol
) {

  hi_df <- openxlsx::read.xlsx(
    xlsxFile = file,
    sheet = sheet,
    colNames = TRUE,
    skipEmptyRows = FALSE,
    skipEmptyCols = FALSE,
    check.names = FALSE
  )

  # Convert every column to numeric
  hi_df[] <- lapply(
    hi_df,
    function(z) {
      suppressWarnings(
        as.numeric(z)
      )
    }
  )

  hi_mat <- as.matrix(hi_df)

  storage.mode(hi_mat) <- "double"

  if (
    nrow(hi_mat) != expected_nrow ||
    ncol(hi_mat) != expected_ncol
  ) {

    stop(
      paste0(
        "Dimension mismatch for sheet '",
        sheet,
        "' in file:\n",
        file,
        "\n\nExpected: ",
        expected_nrow,
        " rows x ",
        expected_ncol,
        " columns\n",
        "Found: ",
        nrow(hi_mat),
        " rows x ",
        ncol(hi_mat),
        " columns."
      )
    )
  }

  hi_mat
}


# ------------------------------------------------------------
# 11. Identify endpoints available in both workbooks
# ------------------------------------------------------------

chronic_endpoints_wbkg <- excel_sheets(
  chronic_hi_wbkg_file
)

chronic_endpoints_nobkg <- excel_sheets(
  chronic_hi_nobkg_file
)

chronic_endpoints <- intersect(
  chronic_endpoints_wbkg,
  chronic_endpoints_nobkg
)

chronic_endpoints

# Check whether either workbook has unmatched sheets
setdiff(
  chronic_endpoints_wbkg,
  chronic_endpoints_nobkg
)

setdiff(
  chronic_endpoints_nobkg,
  chronic_endpoints_wbkg
)


# ------------------------------------------------------------
# 12. Summary function
#
# n is the number of nonmissing grid-cell HI values included
# for that endpoint, scenario, and distance.
# ------------------------------------------------------------

calculate_chronic_hi_distance_summary <- function(
  mat,
  distance_mask,
  endpoint_name,
  scenario_name,
  distance_ft,
  distance_m
) {

  vals <- as.vector(mat)

  if (length(vals) != length(distance_mask)) {
    stop(
      paste0(
        "The HI matrix and spatial distance mask have ",
        "different lengths for endpoint: ",
        endpoint_name
      )
    )
  }

  vals <- vals[distance_mask]
  vals <- vals[is.finite(vals)]

  n_values <- length(vals)

  if (n_values == 0) {

    return(
      data.frame(
        scenario = scenario_name,
        endpoint = endpoint_name,
        distance_ft = distance_ft,
        distance_m = distance_m,
        n = 0,
        min = NA_real_,
        median = NA_real_,
        mean = NA_real_,
        max = NA_real_,
        sd = NA_real_,
        se = NA_real_
      )
    )
  }

  value_sd <- if (n_values > 1) {
    sd(vals)
  } else {
    NA_real_
  }

  value_se <- if (n_values > 1) {
    value_sd / sqrt(n_values)
  } else {
    NA_real_
  }

  data.frame(
    scenario = scenario_name,
    endpoint = endpoint_name,
    distance_ft = distance_ft,
    distance_m = distance_m,
    n = n_values,
    min = min(vals),
    median = median(vals),
    mean = mean(vals),
    max = max(vals),
    sd = value_sd,
    se = value_se
  )
}


# ------------------------------------------------------------
# 13. Read each endpoint and calculate all three scenarios
#
# Scenario definitions:
#
# Without background:
#   HI from chronic_HIs.xlsx
#
# Background only:
#   HI with background - HI without background
#
# Total:
#   HI from chronic_HIs_wbkg.xlsx
#
# Processing one endpoint at a time avoids retaining all
# matrices in memory.
# ------------------------------------------------------------

chronic_hi_distance_summary_list <- list()

summary_counter <- 1

for (ep in chronic_endpoints) {

  message(
    "Processing chronic HI distance summaries for endpoint: ",
    ep
  )

  # Read total / with-background HI
  hi_total_mat <- read_chronic_hi_matrix(
    file = chronic_hi_wbkg_file,
    sheet = ep,
    expected_nrow = chronic_grid_nrow,
    expected_ncol = chronic_grid_ncol
  )

  # Read without-background HI
  hi_nobkg_mat <- read_chronic_hi_matrix(
    file = chronic_hi_nobkg_file,
    sheet = ep,
    expected_nrow = chronic_grid_nrow,
    expected_ncol = chronic_grid_ncol
  )

  stopifnot(
    identical(
      dim(hi_total_mat),
      dim(hi_nobkg_mat)
    )
  )

  # Calculate background-only HI
  hi_background_mat <- hi_total_mat - hi_nobkg_mat

  # Apply the same <25 m mask to every scenario
  hi_total_mat[!chronic_keep_mask] <- NA_real_
  hi_nobkg_mat[!chronic_keep_mask] <- NA_real_
  hi_background_mat[!chronic_keep_mask] <- NA_real_

  # Store the three scenario matrices temporarily
  endpoint_scenario_mats <- list(
    "Without background" = hi_nobkg_mat,
    "Background only" = hi_background_mat,
    "Total" = hi_total_mat
  )

  # Summarize every scenario at every distance
  for (scenario_name in names(endpoint_scenario_mats)) {

    scenario_mat <- endpoint_scenario_mats[[
      scenario_name
    ]]

    for (i in seq_len(nrow(distance_lookup))) {

      current_distance_ft <- distance_lookup$distance_ft[i]
      current_distance_m <- distance_lookup$distance_m[i]

      current_distance_mask <-
        chronic_distance_masks[[
          as.character(current_distance_ft)
        ]]

      chronic_hi_distance_summary_list[[
        summary_counter
      ]] <- calculate_chronic_hi_distance_summary(
        mat = scenario_mat,
        distance_mask = current_distance_mask,
        endpoint_name = ep,
        scenario_name = scenario_name,
        distance_ft = current_distance_ft,
        distance_m = current_distance_m
      )

      summary_counter <- summary_counter + 1
    }
  }

  # Remove temporary matrices before processing next endpoint
  rm(
    hi_total_mat,
    hi_nobkg_mat,
    hi_background_mat,
    endpoint_scenario_mats
  )
}


# ------------------------------------------------------------
# 14. Combine all results
# ------------------------------------------------------------

chronic_hi_distance_summary <- bind_rows(
  chronic_hi_distance_summary_list
) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Without background",
        "Background only",
        "Total"
      )
    )
  ) %>%
  arrange(
    endpoint,
    distance_ft,
    scenario
  )

chronic_hi_distance_summary
```





#checks and saving
```{r}

# ------------------------------------------------------------
# 15. Validate expected number of rows
#
# Expected:
#   number of endpoints
#   x 10 distances
#   x 3 scenarios
# ------------------------------------------------------------

expected_summary_rows <-
  length(chronic_endpoints) *
  nrow(distance_lookup) *
  3

actual_summary_rows <-
  nrow(chronic_hi_distance_summary)

data.frame(
  expected_rows = expected_summary_rows,
  actual_rows = actual_summary_rows,
  rows_match = expected_summary_rows == actual_summary_rows
)

stopifnot(
  expected_summary_rows == actual_summary_rows
)


# ------------------------------------------------------------
# 16. Additional checks
# ------------------------------------------------------------

# Number of rows per endpoint
chronic_hi_distance_summary %>%
  count(endpoint)

# Number of rows per scenario
chronic_hi_distance_summary %>%
  count(scenario)

# Number of rows per distance
chronic_hi_distance_summary %>%
  count(distance_ft)


# ------------------------------------------------------------
# 17. Save combined long-format CSV
# ------------------------------------------------------------

write.csv(
  chronic_hi_distance_summary,
  file = chronic_hi_distance_output_csv,
  row.names = FALSE
)


# ------------------------------------------------------------
# 18. Create a wider reporting version
# ------------------------------------------------------------

chronic_hi_distance_summary_wide <-
  chronic_hi_distance_summary %>%
  mutate(
    scenario = as.character(scenario),
    scenario = recode(
      scenario,
      "Without background" = "without_background",
      "Background only" = "background_only",
      "Total" = "total"
    )
  ) %>%
  pivot_wider(
    id_cols = c(
      endpoint,
      distance_ft,
      distance_m
    ),
    names_from = scenario,
    values_from = c(
      n,
      min,
      median,
      mean,
      max,
      sd,
      se
    ),
    names_glue = "{scenario}_{.value}"
  ) %>%
  select(
    endpoint,
    distance_ft,
    distance_m,

    # Without background
    without_background_n,
    without_background_min,
    without_background_median,
    without_background_mean,
    without_background_max,
    without_background_sd,
    without_background_se,

    # Background only
    background_only_n,
    background_only_min,
    background_only_median,
    background_only_mean,
    background_only_max,
    background_only_sd,
    background_only_se,

    # Total
    total_n,
    total_min,
    total_median,
    total_mean,
    total_max,
    total_sd,
    total_se
  ) %>%
  arrange(
    endpoint,
    distance_ft
  )

chronic_hi_distance_summary_wide


# ------------------------------------------------------------
# 19. Save an Excel workbook with long and wide tables
# ------------------------------------------------------------

chronic_summary_wb <- createWorkbook()

addWorksheet(
  chronic_summary_wb,
  "Long summary"
)

writeData(
  chronic_summary_wb,
  sheet = "Long summary",
  x = chronic_hi_distance_summary
)

addWorksheet(
  chronic_summary_wb,
  "Wide summary"
)

writeData(
  chronic_summary_wb,
  sheet = "Wide summary",
  x = chronic_hi_distance_summary_wide
)

addWorksheet(
  chronic_summary_wb,
  "Distance grid counts"
)

writeData(
  chronic_summary_wb,
  sheet = "Distance grid counts",
  x = distance_grid_counts
)

# Optional formatting
freezePane(
  chronic_summary_wb,
  sheet = "Long summary",
  firstRow = TRUE
)

freezePane(
  chronic_summary_wb,
  sheet = "Wide summary",
  firstRow = TRUE
)

freezePane(
  chronic_summary_wb,
  sheet = "Distance grid counts",
  firstRow = TRUE
)

setColWidths(
  chronic_summary_wb,
  sheet = "Long summary",
  cols = 1:ncol(chronic_hi_distance_summary),
  widths = "auto"
)

setColWidths(
  chronic_summary_wb,
  sheet = "Wide summary",
  cols = 1:ncol(chronic_hi_distance_summary_wide),
  widths = "auto"
)

setColWidths(
  chronic_summary_wb,
  sheet = "Distance grid counts",
  cols = 1:ncol(distance_grid_counts),
  widths = "auto"
)

saveWorkbook(
  chronic_summary_wb,
  file = chronic_hi_distance_output_xlsx,
  overwrite = TRUE
)
```






### cancer analysis ###




#fix chemical names between iurs and data
```{r}
# check chemical names
unique(iurs$Chemical)
names(TWA_list)

# fix mismatches 
iurs$Chemical[iurs$Chemical == "Benzene"] <- "benzene"
iurs$Chemical[iurs$Chemical == "Ethylbenzene"] <- "ethylbenzene"
iurs$Chemical[iurs$Chemical == "C2HCl3"] <- "c2hcl3"
iurs$Chemical[iurs$Chemical == "C2Cl4"] <- "c2cl4"
```


#calculate excess cancer risk, assuming 33 years of exposure (3 years of data in pre-production and 30 in production)
```{r}
CancerRisk_list <- list()

# exposure duration adjustment
ED_adjust <- 33 / 78

#formula
for (chem in names(TWA_list)) {
  
  # get IUR for this chemical
  iur_val <- iurs %>%
    dplyr::filter(Chemical == chem) %>%
    dplyr::pull(IUR..ppb.)
  
  # skip if no IUR
  if (length(iur_val) == 0) next
  
  iur_val <- iur_val[1]
  
  # calculate risk with exposure duration adjustment
  CancerRisk_list[[chem]] <- TWA_list[[chem]] * iur_val * ED_adjust
}

#sum cancer risk across chemicals
CancerRisk_total <- Reduce(`+`, CancerRisk_list)

#apply the <25 m mask
CancerRisk_total[!mask] <- NA

#checks
# dimensions
dim(CancerRisk_total)
dim(TWA_list[[1]])

# summary
summary(as.vector(CancerRisk_total))

# inspect one chemical contribution
summary(as.vector(CancerRisk_list[[1]]))

## save cancer risk data for each chemical separated by tabs into one excel file for reporting ##

output_file <- "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_risk_wbkg.xlsx"

wb <- createWorkbook()

for (chem in names(CancerRisk_list)) {

  # skip missing/empty entries (safety)
  if (is.null(CancerRisk_list[[chem]])) next

  sheet_name <- substr(chem, 1, 31)

  addWorksheet(wb, sheet_name)

  chem_df <- as.data.frame(CancerRisk_list[[chem]])

  writeData(
    wb,
    sheet = sheet_name,
    x = chem_df
  )
}

saveWorkbook(
  wb,
  file = output_file,
  overwrite = TRUE
)

```


#summary stats
```{r}
#total cancer risk
cancer_summary_tot <- calc_summary(CancerRisk_total)

cancer_summary_tot

write.csv(
  cancer_summary_tot,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_tot.csv",
  row.names = FALSE
)

#summary per chemical
cancer_summary_ind <- do.call(
  rbind,
  lapply(names(CancerRisk_list), function(chem) {
    cbind(
      chemical = chem,
      calc_summary(CancerRisk_list[[chem]])
    )
  })
)

# clean row names
rownames(cancer_summary_ind) <- NULL

cancer_summary_ind

write.csv(
  cancer_summary_ind,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_ind_wbkg.csv",
  row.names = FALSE
)
```


#map cancer risks
```{r}
#define bounds
x_min <- 495300
x_max <- 501700

y_min <- 4424300
y_max <- 4427200

make_df <- function(mat, x, y) {
  data.frame(
    x = as.vector(x),
    y = as.vector(y),
    Risk = as.vector(mat)
  )
}

# read well pad coordinates
wellpad_cords <- read.csv(
  "~/CO_CRA/CO_CRA_aggregate/data/broomfield/wellpad_cords.csv"
)

# create 610 m dissolved/unioned setback boundary
radius <- 610

wellpad_sf <- st_as_sf(
  wellpad_cords,
  coords = c("X", "Y"),
  crs = NA
)

setback_boundary <- wellpad_sf %>%
  st_buffer(dist = radius) %>%
  st_union() %>%
  st_boundary()

setback_boundary_df <- setback_boundary %>%
  st_as_sf() %>%
  mutate(id = row_number()) %>%
  st_cast("LINESTRING") %>%
  st_coordinates() %>%
  as.data.frame()

# fixed/define legend
risk_min <- 1.5e-4
risk_max <- 5.5e-4

risk_breaks <- c(1.5e-4, 2.5e-4, 3.5e-4, 4.5e-4, 5.5e-4)

contour_levels <- c(1e-6, 1e-4)

# make data
df <- make_df(CancerRisk_total, x, y)

# plot
plot_cancer_tot <- ggplot(df, aes(x = x, y = y, fill = Risk)) +

  geom_raster() +

  # outer 610 m setback boundary only
  geom_path(
    data = setback_boundary_df,
    aes(x = X, y = Y, group = L1),
    color = "black",
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +

scale_fill_viridis_c(
  name = "Cancer Risk",
  trans = "log10",
  limits = c(risk_min, risk_max),
  breaks = risk_breaks,
  labels = scales::label_scientific(digits = 2),
  oob = scales::squish
) +

  geom_contour(
    aes(z = Risk),
    breaks = contour_levels,
    color = "red",
    linewidth = 0.7
  ) +

  coord_equal(
    xlim = c(x_min, x_max),
    ylim = c(y_min, y_max),
    expand = FALSE
  ) +

  labs(
    title = "Total Excess Cancer Risk",
    x = "X (m)",
    y = "Y (m)"
  ) +

  theme_minimal()

plot_cancer_tot

ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_tot_wbkg.png",
  plot = plot_cancer_tot,
  width = 6,
  height = 6,
  dpi = 300
)
```




## map of just background cancer risk ## 



#load cancer risks that were calculated without background
```{r}
CancerRisk_total_nobkg <- readRDS(
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/CancerRisk_total_nobkg.rds"
)

CancerRisk_list_nobkg <- readRDS(
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/CancerRisk_list_nobkg.rds"
)

#check dimensions
dim(CancerRisk_total)
dim(CancerRisk_total_nobkg)

all(dim(CancerRisk_total) == dim(CancerRisk_total_nobkg))
```


#calculate background-only cancer risks (cancer risks wbkg - cancer risks w/o bkg)
```{r}
#total
CancerRisk_background <-
  CancerRisk_total -
  CancerRisk_total_nobkg

CancerRisk_background[!mask] <- NA

#by chemical
CancerRisk_background_list <- list()

common_chems <- intersect(
  names(CancerRisk_list),
  names(CancerRisk_list_nobkg)
)

for (chem in common_chems) {

  CancerRisk_background_list[[chem]] <-
    CancerRisk_list[[chem]] -
    CancerRisk_list_nobkg[[chem]]
}

#quick check
summary(as.vector(CancerRisk_background))

lapply(
  CancerRisk_background_list,
  function(x) summary(as.vector(x))
)

## save risks into one excel file with chemicals separated by tabs for reporting ##

output_file <- "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_risk_backgroundonly.xlsx"

wb <- createWorkbook()

for (chem in names(CancerRisk_background_list)) {

  # skip missing entries (safety)
  if (is.null(CancerRisk_background_list[[chem]])) next

  sheet_name <- substr(chem, 1, 31)

  addWorksheet(wb, sheet_name)

  chem_df <- as.data.frame(CancerRisk_background_list[[chem]])

  writeData(
    wb,
    sheet = sheet_name,
    x = chem_df
  )
}

saveWorkbook(
  wb,
  file = output_file,
  overwrite = TRUE
)
```


#summary of background cancer risks
```{r}
# total background cancer risk
cancer_summary_background_tot <- calc_summary(CancerRisk_background)

cancer_summary_background_tot

write.csv(
  cancer_summary_background_tot,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_background_tot.csv",
  row.names = FALSE
)

# summary per chemical
cancer_summary_background_ind <- do.call(
  rbind,
  lapply(names(CancerRisk_background_list), function(chem) {

    cbind(
      chemical = chem,
      calc_summary(CancerRisk_background_list[[chem]])
    )

  })
)

# clean row names
rownames(cancer_summary_background_ind) <- NULL

cancer_summary_background_ind

write.csv(
  cancer_summary_background_ind,
  file = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_background_ind.csv",
  row.names = FALSE
)
```


#map background cancer risks
```{r}
make_df <- function(mat, x, y) {
  data.frame(
    x = as.vector(x),
    y = as.vector(y),
    Risk = as.vector(mat)
  )
}

# read well pad coordinates if not already loaded
wellpad_cords <- read.csv(
  "~/CO_CRA/CO_CRA_aggregate/data/broomfield/wellpad_cords.csv"
)

# create outer-only 610 m setback boundary
radius <- 610

wellpad_sf <- st_as_sf(
  wellpad_cords,
  coords = c("X", "Y"),
  crs = NA
)

setback_boundary <- wellpad_sf %>%
  st_buffer(dist = radius) %>%
  st_union() %>%
  st_boundary()

setback_boundary_df <- setback_boundary %>%
  st_as_sf() %>%
  st_cast("LINESTRING") %>%
  st_coordinates() %>%
  as.data.frame()

# fixed legend
risk_min <- 1.5e-4
risk_max <- 5.5e-4

risk_breaks <- c(1.5e-4, 2.5e-4, 3.5e-4, 4.5e-4, 5.5e-4)

# regulatory contour limits
contour_levels <- c(1e-6, 1e-4)

# create dataframe from BACKGROUND risk
df <- make_df(CancerRisk_background, x, y)

# plot
plot_cancer_background <- ggplot(
  df,
  aes(x = x, y = y, fill = Risk)
) +
  geom_raster() +

scale_fill_viridis_c(
  name = "Cancer Risk",
  trans = "log10",
  limits = c(risk_min, risk_max),
  breaks = risk_breaks,
  labels = scales::label_scientific(digits = 2),
  oob = scales::squish
) +

  geom_contour(
    aes(z = Risk),
    breaks = contour_levels,
    color = "red",
    linewidth = 0.7
  ) +

  # outer-only 610 m setback boundary
  geom_path(
    data = setback_boundary_df,
    aes(x = X, y = Y, group = L1),
    color = "black",
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +

  coord_equal(
    xlim = c(x_min, x_max),
    ylim = c(y_min, y_max),
    expand = FALSE
  ) +

  labs(
    title = "Background Excess Cancer Risk",
    x = "X (m)",
    y = "Y (m)"
  ) +

  theme_minimal()

# view
plot_cancer_background

# save
ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_background.png",
  plot = plot_cancer_background,
  width = 6,
  height = 6,
  dpi = 300
)
```



#panel together total cancer and background cancer graphs
```{r}
# -----------------------------
# Scientific notation labels for cancer risk legend
# -----------------------------

risk_legend_labels <- c(
  expression(1.5 %*% 10^{-4}),
  expression(2.5 %*% 10^{-4}),
  expression(3.5 %*% 10^{-4}),
  expression(4.5 %*% 10^{-4}),
  expression(5.5 %*% 10^{-4})
)


# -----------------------------
# A. Total cancer risk panel
# -----------------------------

plot_cancer_tot_panel <- plot_cancer_tot +

  labs(
    title = NULL
  ) +

  # Add commas to X-axis coordinates
  scale_x_continuous(
    labels = scales::label_comma()
  ) +

  # Add commas to Y-axis coordinates
  scale_y_continuous(
    labels = scales::label_comma()
  ) +

  # Re-define fill scale so legend uses:
  # 1.5 × 10^-4 rather than 1.5e-04
  scale_fill_viridis_c(
    name = "Cancer Risk",
    trans = "log10",
    limits = c(risk_min, risk_max),
    breaks = risk_breaks,
    labels = risk_legend_labels,
    oob = scales::squish
  ) +

  theme(
    legend.position = "right",
    plot.margin = margin(2, 2, 2, 2)
  )


# -----------------------------
# B. Background cancer risk panel
# -----------------------------

plot_cancer_background_panel <- plot_cancer_background +

  labs(
    title = NULL
  ) +

  # Add commas to X-axis coordinates
  scale_x_continuous(
    labels = scales::label_comma()
  ) +

  # Add commas to Y-axis coordinates
  scale_y_continuous(
    labels = scales::label_comma()
  ) +

  # Re-define fill scale so legend uses:
  # 1.5 × 10^-4 rather than 1.5e-04
  scale_fill_viridis_c(
    name = "Cancer Risk",
    trans = "log10",
    limits = c(risk_min, risk_max),
    breaks = risk_breaks,
    labels = risk_legend_labels,
    oob = scales::squish
  ) +

  theme(
    legend.position = "right",
    plot.margin = margin(2, 2, 2, 2)
  )


# -----------------------------
# Combine panels
# -----------------------------

combined_cancer_panel <- plot_cancer_tot_panel /
  plot_cancer_background_panel +

  plot_layout(
    ncol = 1,
    guides = "collect"
  ) +

  plot_annotation(
    tag_levels = "A"
  ) &

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    plot.tag = element_text(
      size = 14,
      face = "bold"
    )
  )


# View
combined_cancer_panel


# -----------------------------
# Save
# -----------------------------

ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_total_background_panel.png",
  plot = combined_cancer_panel,
  width = 7.5,
  height = 6.5,
  dpi = 300
)
```



### getting range of cancer risk per chemical (and total) at 2,000 ft ###



#filter for grid cells that overlap with 610 m
```{r}
radius <- 610
grid_res <- 50
half_res <- grid_res / 2

wellpad_cords <- read.csv(
  "~/CO_CRA/CO_CRA_aggregate/data/broomfield/wellpad_cords.csv"
)

grid_cells <- data.frame(
  cell_id = seq_along(as.vector(x)),
  x = as.vector(x),
  y = as.vector(y)
) %>%
  rowwise() %>%
  mutate(
    geometry = list(
      st_polygon(list(matrix(
        c(
          x - half_res, y - half_res,
          x + half_res, y - half_res,
          x + half_res, y + half_res,
          x - half_res, y + half_res,
          x - half_res, y - half_res
        ),
        ncol = 2,
        byrow = TRUE
      )))
    )
  ) %>%
  ungroup() %>%
  st_as_sf(crs = NA)

wellpad_sf <- st_as_sf(
  wellpad_cords,
  coords = c("X", "Y"),
  crs = NA
)

circle_lines <- wellpad_sf %>%
  st_buffer(dist = radius) %>%
  st_boundary()

grid_cells$at_610m <- lengths(st_intersects(grid_cells, circle_lines)) > 0

sum(grid_cells$at_610m)
```


#calculate cancer risk summaries w/ and w/o background
```{r}
# function to summarize cancer risks at 610 m
calc_cancer_summary_at_610m <- function(mat, at_610m, chemical_name, scenario_name) {

  vals <- as.vector(mat)
  vals <- vals[at_610m]
  vals <- vals[!is.na(vals)]

  data.frame(
    scenario = scenario_name,
    chemical = chemical_name,
    n_grids = length(vals),
    min_risk = min(vals),
    median_risk = median(vals),
    mean_risk = mean(vals),
    max_risk = max(vals)
  )
}

# wrapper for individual chemicals + total
summarize_cancer_at_610m <- function(risk_list, total_risk, at_610m, scenario_name) {

  individual_summary <- do.call(
    rbind,
    lapply(names(risk_list), function(chem) {

      calc_cancer_summary_at_610m(
        mat = risk_list[[chem]],
        at_610m = at_610m,
        chemical_name = chem,
        scenario_name = scenario_name
      )
    })
  )

  total_summary <- calc_cancer_summary_at_610m(
    mat = total_risk,
    at_610m = at_610m,
    chemical_name = "Total cancer risk",
    scenario_name = scenario_name
  )

  rbind(individual_summary, total_summary)
}

# cancer risks WITHOUT background at 610 m
cancer_610m_nobkg_summary <- summarize_cancer_at_610m(
  risk_list = CancerRisk_list_nobkg,
  total_risk = CancerRisk_total_nobkg,
  at_610m = grid_cells$at_610m,
  scenario_name = "Without background"
)

# cancer risks WITH background at 610 m
cancer_610m_wbkg_summary <- summarize_cancer_at_610m(
  risk_list = CancerRisk_list,
  total_risk = CancerRisk_total,
  at_610m = grid_cells$at_610m,
  scenario_name = "With background"
)

# combine
cancer_610m_summary <- bind_rows(
  cancer_610m_nobkg_summary,
  cancer_610m_wbkg_summary
)

cancer_610m_summary

write.csv(
  cancer_610m_summary,
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_at_2000ft_nobkg_wbkg.csv",
  row.names = FALSE
)
```




#figure of average individual chemical contributes to cancer risks at 2000 ft, with the x axis being the different scenarios: 1) without background, 2) background only, 3) total
```{r}
# -----------------------------
# Chemical labels and colors
# -----------------------------

chem_labels <- c(
  "benzene" = "Benzene",
  "ethylbenzene" = "Ethylbenzene",
  "c2cl4" = "PCE",
  "c2hcl3" = "TCE"
)

chem_colors <- c(
  "Benzene"      = "#D9E57A",
  "Ethylbenzene" = "#C7E9B4",
  "PCE"          = "#41B6C4",
  "TCE"          = "#225EA8"
)


# -----------------------------
# 1. Prepare no-background and with-background data
# -----------------------------

risk_610m_plot_df <- cancer_610m_summary %>%
  filter(chemical != "Total cancer risk") %>%
  filter(scenario %in% c("Without background", "With background")) %>%
  mutate(
    chemical = recode(
      chemical,
      !!!chem_labels
    ),
    scenario = factor(
      scenario,
      levels = c(
        "Without background",
        "Background only",
        "With background"
      ),
      labels = c(
        "UOG Only",
        "Background Only",
        "Total"
      )
    )
  ) %>%
  select(
    scenario,
    chemical,
    mean_risk
  )


# -----------------------------
# 2. Prepare background-only data
# -----------------------------

background_plot_df <- cancer_summary_background_ind %>%
  mutate(
    scenario = "Background only",
    chemical = recode(
      chemical,
      !!!chem_labels
    ),
    mean_risk = mean
  ) %>%
  select(
    scenario,
    chemical,
    mean_risk
  ) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Without background",
        "Background only",
        "With background"
      ),
      labels = c(
        "UOG Only",
        "Background Only",
        "Total"
      )
    )
  )


# -----------------------------
# 3. Combine
# -----------------------------

risk_stacked_plot_df <- bind_rows(
  risk_610m_plot_df,
  background_plot_df
) %>%
  filter(
    chemical %in% names(chem_colors)
  ) %>%
  mutate(
    chemical = factor(
      chemical,
      levels = c(
        "Benzene",
        "Ethylbenzene",
        "PCE",
        "TCE"
      )
    )
  )


# Check values
risk_stacked_plot_df


# -----------------------------
# 4. Plot
# -----------------------------

cancer_risk_stacked_bar <- ggplot(
  risk_stacked_plot_df,
  aes(
    x = scenario,
    y = mean_risk,
    fill = chemical
  )
) +

  geom_col(
    width = 0.7,
    color = "black",
    linewidth = 0.2
  ) +

  # Cancer risk reference lines
  geom_hline(
    yintercept = c(1e-6, 1e-4),
    linetype = "dashed",
    color = "red",
    linewidth = 0.8
  ) +

  # Chemical colors and legend labels
  scale_fill_manual(
    name = "Chemical",
    values = chem_colors,
    labels = c(
      "Benzene" = "Benzene (B)",
      "Ethylbenzene" = "Ethylbenzene (E)",
      "PCE" = "PCE (P)",
      "TCE" = "TCE (T)"
    )
  ) +

  # Y axis using scientific notation:
  # e.g., 1 × 10^-6 instead of 1.0e-06
  scale_y_continuous(
    limits = c(0, 2.6e-4),
    breaks = c(
      1e-6,
      5e-5,
      1e-4,
      1.5e-4,
      2e-4,
      2.5e-4
    ),
    labels = c(
      expression(1 %*% 10^{-6}),
      expression(5 %*% 10^{-5}),
      expression(1 %*% 10^{-4}),
      expression(1.5 %*% 10^{-4}),
      expression(2 %*% 10^{-4}),
      expression(2.5 %*% 10^{-4})
    ),
    expand = c(0, 0)
  ) +

  labs(
    x = NULL,
    y = "Average Cancer Risk at 2000 ft (n = 553 grids)"
  ) +

  theme_minimal() +

  theme(
    legend.position = "right",
    axis.text.x = element_text(
      angle = 0,
      hjust = 0.5,
      size = 11
    )
  )


# Display plot
cancer_risk_stacked_bar


# -----------------------------
# 5. Save plot
# -----------------------------

ggsave(
  filename = "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_mean_risk_stacked_bar.png",
  plot = cancer_risk_stacked_bar,
  width = 7,
  height = 5,
  dpi = 300
)
```





### cancer risk summary table for individual risks and cumulative risks at all distances, for UOGD, background, total ###


#load and check spatial analysis from the chronic workflow, and calculate the summaries
```{r}
# ------------------------------------------------------------
# 1. Define output files
# ------------------------------------------------------------

cancer_distance_output_csv <-
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_500_to_5000ft_all_scenarios.csv"

cancer_distance_output_xlsx <-
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/cancer_summary_500_to_5000ft_all_scenarios.xlsx"


# ------------------------------------------------------------
# 2. Confirm that the required spatial objects exist
#
# These objects were created in the chronic HI distance
# workflow and are reused here without modification.
# ------------------------------------------------------------

required_spatial_objects <- c(
  "distance_lookup",
  "chronic_distance_masks",
  "chronic_keep_mask",
  "chronic_grid_nrow",
  "chronic_grid_ncol",
  "distance_grid_counts"
)

missing_spatial_objects <- required_spatial_objects[
  !vapply(
    required_spatial_objects,
    exists,
    logical(1),
    inherits = TRUE
  )
]

if (length(missing_spatial_objects) > 0) {

  stop(
    paste0(
      "Run Steps 1–9 of the chronic distance workflow first. ",
      "The following required objects are missing: ",
      paste(missing_spatial_objects, collapse = ", ")
    )
  )
}


# ------------------------------------------------------------
# 3. Create separate copies of the with-background cancer risks
#
# These copies prevent this workflow from modifying the
# existing CancerRisk_list and CancerRisk_total objects.
# ------------------------------------------------------------

cancer_wbkg_list <- lapply(
  CancerRisk_list,
  function(mat) {

    copied_mat <- mat
    storage.mode(copied_mat) <- "double"

    copied_mat
  }
)

cancer_wbkg_total <- CancerRisk_total
storage.mode(cancer_wbkg_total) <- "double"


# ------------------------------------------------------------
# 4. Load the previously saved without-background cancer risks
#
# These files are already created by the existing no-background
# cancer analysis.
# ------------------------------------------------------------

cancer_nobkg_list <- readRDS(
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/CancerRisk_list_nobkg.rds"
)

cancer_nobkg_total <- readRDS(
  "~/CO_CRA/CO_CRA_aggregate/output/chronic/CancerRisk_total_nobkg.rds"
)

# Ensure numeric matrix storage
cancer_nobkg_list <- lapply(
  cancer_nobkg_list,
  function(mat) {

    storage.mode(mat) <- "double"

    mat
  }
)

storage.mode(cancer_nobkg_total) <- "double"


# ------------------------------------------------------------
# 5. Check that the chemical names match between scenarios
# ------------------------------------------------------------

cancer_chems_wbkg <- names(cancer_wbkg_list)
cancer_chems_nobkg <- names(cancer_nobkg_list)

# These should both return character(0)
setdiff(
  cancer_chems_wbkg,
  cancer_chems_nobkg
)

setdiff(
  cancer_chems_nobkg,
  cancer_chems_wbkg
)

cancer_common_chems <- intersect(
  cancer_chems_wbkg,
  cancer_chems_nobkg
)

if (length(cancer_common_chems) == 0) {

  stop(
    paste0(
      "No matching chemicals were found between the ",
      "with-background and without-background cancer-risk lists."
    )
  )
}


# ------------------------------------------------------------
# 6. Check matrix dimensions
# ------------------------------------------------------------

expected_cancer_dimensions <- c(
  chronic_grid_nrow,
  chronic_grid_ncol
)

for (chem in cancer_common_chems) {

  if (
    !identical(
      dim(cancer_wbkg_list[[chem]]),
      expected_cancer_dimensions
    )
  ) {

    stop(
      paste0(
        "Dimension mismatch for with-background cancer risk: ",
        chem
      )
    )
  }

  if (
    !identical(
      dim(cancer_nobkg_list[[chem]]),
      expected_cancer_dimensions
    )
  ) {

    stop(
      paste0(
        "Dimension mismatch for without-background cancer risk: ",
        chem
      )
    )
  }

  if (
    !identical(
      dim(cancer_wbkg_list[[chem]]),
      dim(cancer_nobkg_list[[chem]])
    )
  ) {

    stop(
      paste0(
        "The with-background and without-background matrices ",
        "do not match for chemical: ",
        chem
      )
    )
  }
}

stopifnot(
  identical(
    dim(cancer_wbkg_total),
    expected_cancer_dimensions
  )
)

stopifnot(
  identical(
    dim(cancer_nobkg_total),
    expected_cancer_dimensions
  )
)

stopifnot(
  identical(
    dim(cancer_wbkg_total),
    dim(cancer_nobkg_total)
  )
)


# ------------------------------------------------------------
# 7. Calculate background-only individual cancer risks
#
# Background only =
#   cancer risk with background
#   minus cancer risk without background
# ------------------------------------------------------------

cancer_background_list <- setNames(
  vector(
    mode = "list",
    length = length(cancer_common_chems)
  ),
  cancer_common_chems
)

for (chem in cancer_common_chems) {

  cancer_background_list[[chem]] <-
    cancer_wbkg_list[[chem]] -
    cancer_nobkg_list[[chem]]
}


# ------------------------------------------------------------
# 8. Calculate total cancer risk for each scenario
#
# The individual risk matrices are summed cell by cell.
#
# This provides an independent check against the existing
# total-risk matrices.
# ------------------------------------------------------------

cancer_wbkg_total_from_individual <- Reduce(
  `+`,
  cancer_wbkg_list[cancer_common_chems]
)

cancer_nobkg_total_from_individual <- Reduce(
  `+`,
  cancer_nobkg_list[cancer_common_chems]
)

cancer_background_total <- Reduce(
  `+`,
  cancer_background_list[cancer_common_chems]
)


# ------------------------------------------------------------
# 9. Check existing totals against sums of individual risks
#
# These should return TRUE or have extremely small numerical
# differences caused only by floating-point precision.
# ------------------------------------------------------------

isTRUE(
  all.equal(
    cancer_wbkg_total,
    cancer_wbkg_total_from_individual,
    check.attributes = FALSE
  )
)

isTRUE(
  all.equal(
    cancer_nobkg_total,
    cancer_nobkg_total_from_individual,
    check.attributes = FALSE
  )
)

# Check the maximum absolute difference
max(
  abs(
    cancer_wbkg_total -
    cancer_wbkg_total_from_individual
  ),
  na.rm = TRUE
)

max(
  abs(
    cancer_nobkg_total -
    cancer_nobkg_total_from_individual
  ),
  na.rm = TRUE
)


# ------------------------------------------------------------
# 10. Use the summed individual risks as the scenario totals
#
# This guarantees that "Total cancer risk" is explicitly the
# sum of all individual cancer risks included in the table.
# ------------------------------------------------------------

cancer_wbkg_total_summary_mat <-
  cancer_wbkg_total_from_individual

cancer_nobkg_total_summary_mat <-
  cancer_nobkg_total_from_individual

cancer_background_total_summary_mat <-
  cancer_background_total


# ------------------------------------------------------------
# 11. Apply the <25 m exclusion mask to copies only
# ------------------------------------------------------------

for (chem in cancer_common_chems) {

  cancer_wbkg_list[[chem]][
    !chronic_keep_mask
  ] <- NA_real_

  cancer_nobkg_list[[chem]][
    !chronic_keep_mask
  ] <- NA_real_

  cancer_background_list[[chem]][
    !chronic_keep_mask
  ] <- NA_real_
}

cancer_wbkg_total_summary_mat[
  !chronic_keep_mask
] <- NA_real_

cancer_nobkg_total_summary_mat[
  !chronic_keep_mask
] <- NA_real_

cancer_background_total_summary_mat[
  !chronic_keep_mask
] <- NA_real_


# ------------------------------------------------------------
# 14. Add total cancer risk to each scenario list
#
# Individual chemicals and total risk can now be processed
# through the same summary function.
# ------------------------------------------------------------

cancer_wbkg_summary_list <- cancer_wbkg_list[
  cancer_common_chems
]

cancer_wbkg_summary_list[[
  "Total cancer risk"
]] <- cancer_wbkg_total_summary_mat


cancer_nobkg_summary_list <- cancer_nobkg_list[
  cancer_common_chems
]

cancer_nobkg_summary_list[[
  "Total cancer risk"
]] <- cancer_nobkg_total_summary_mat


cancer_background_summary_list <-
  cancer_background_list[
    cancer_common_chems
  ]

cancer_background_summary_list[[
  "Total cancer risk"
]] <- cancer_background_total_summary_mat


# ------------------------------------------------------------
# 15. Store the three scenarios in one nested list
# ------------------------------------------------------------

cancer_scenario_lists <- list(
  "Without background" = cancer_nobkg_summary_list,
  "Background only" = cancer_background_summary_list,
  "Total" = cancer_wbkg_summary_list
)


# ------------------------------------------------------------
# 16. Function to calculate cancer-risk summary statistics
#
# n is the number of finite grid-cell values included for that
# chemical/total, scenario, and distance.
# ------------------------------------------------------------

calculate_cancer_distance_summary <- function(
  mat,
  distance_mask,
  risk_name,
  scenario_name,
  distance_ft,
  distance_m
) {

  vals <- as.vector(mat)

  if (length(vals) != length(distance_mask)) {

    stop(
      paste0(
        "The cancer-risk matrix and spatial distance mask ",
        "have different lengths for: ",
        risk_name
      )
    )
  }

  vals <- vals[distance_mask]
  vals <- vals[is.finite(vals)]

  n_values <- length(vals)

  if (n_values == 0) {

    return(
      data.frame(
        scenario = scenario_name,
        risk_type = risk_name,
        distance_ft = distance_ft,
        distance_m = distance_m,
        n = 0,
        min = NA_real_,
        median = NA_real_,
        mean = NA_real_,
        max = NA_real_,
        sd = NA_real_,
        se = NA_real_
      )
    )
  }

  value_sd <- if (n_values > 1) {
    sd(vals)
  } else {
    NA_real_
  }

  value_se <- if (n_values > 1) {
    value_sd / sqrt(n_values)
  } else {
    NA_real_
  }

  data.frame(
    scenario = scenario_name,
    risk_type = risk_name,
    distance_ft = distance_ft,
    distance_m = distance_m,
    n = n_values,
    min = min(vals),
    median = median(vals),
    mean = mean(vals),
    max = max(vals),
    sd = value_sd,
    se = value_se
  )
}


# ------------------------------------------------------------
# 17. Calculate summaries for every:
#
# scenario
# × individual chemical or total
# × distance
# ------------------------------------------------------------

cancer_distance_summary_list <- list()

cancer_summary_counter <- 1

for (
  scenario_name in names(cancer_scenario_lists)
) {

  current_scenario_list <-
    cancer_scenario_lists[[scenario_name]]

  for (
    risk_name in names(current_scenario_list)
  ) {

    message(
      "Processing: ",
      scenario_name,
      " | ",
      risk_name
    )

    current_risk_mat <-
      current_scenario_list[[risk_name]]

    for (i in seq_len(nrow(distance_lookup))) {

      current_distance_ft <-
        distance_lookup$distance_ft[i]

      current_distance_m <-
        distance_lookup$distance_m[i]

      current_distance_mask <-
        chronic_distance_masks[[
          as.character(current_distance_ft)
        ]]

      cancer_distance_summary_list[[
        cancer_summary_counter
      ]] <- calculate_cancer_distance_summary(
        mat = current_risk_mat,
        distance_mask = current_distance_mask,
        risk_name = risk_name,
        scenario_name = scenario_name,
        distance_ft = current_distance_ft,
        distance_m = current_distance_m
      )

      cancer_summary_counter <-
        cancer_summary_counter + 1
    }
  }
}


# ------------------------------------------------------------
# 18. Combine all results
# ------------------------------------------------------------

cancer_distance_summary <- bind_rows(
  cancer_distance_summary_list
) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Without background",
        "Background only",
        "Total"
      )
    )
  ) %>%
  arrange(
    risk_type,
    distance_ft,
    scenario
  )

cancer_distance_summary
```




#checks and saving
```{r}
# ------------------------------------------------------------
# 19. Validate the expected number of rows
#
# Expected:
#   number of individual chemicals plus total cancer risk
#   x 10 distances
#   x 3 scenarios
# ------------------------------------------------------------

number_of_risk_types <-
  length(cancer_common_chems) + 1

expected_cancer_summary_rows <-
  number_of_risk_types *
  nrow(distance_lookup) *
  3

actual_cancer_summary_rows <-
  nrow(cancer_distance_summary)

data.frame(
  expected_rows = expected_cancer_summary_rows,
  actual_rows = actual_cancer_summary_rows,
  rows_match =
    expected_cancer_summary_rows ==
    actual_cancer_summary_rows
)

stopifnot(
  expected_cancer_summary_rows ==
  actual_cancer_summary_rows
)


# ------------------------------------------------------------
# 20. Additional checks
# ------------------------------------------------------------

# Number of rows per risk type
cancer_distance_summary %>%
  count(risk_type)

# Number of rows per scenario
cancer_distance_summary %>%
  count(scenario)

# Number of rows per distance
cancer_distance_summary %>%
  count(distance_ft)

# Check n against spatial grid counts
cancer_distance_summary %>%
  distinct(
    risk_type,
    scenario,
    distance_ft,
    n
  ) %>%
  left_join(
    distance_grid_counts,
    by = "distance_ft"
  )


# ------------------------------------------------------------
# 21. Optional chemical-name labels
#
# Add any other chemicals with cancer risk values as needed.
# Unlisted chemical names remain unchanged.
# ------------------------------------------------------------

cancer_risk_labels <- c(
  "benzene" = "Benzene",
  "ethylbenzene" = "Ethylbenzene",
  "c2cl4" = "PCE",
  "c2hcl3" = "TCE",
  "Total cancer risk" = "Total cancer risk"
)

cancer_distance_summary <- cancer_distance_summary %>%
  mutate(
    risk_type = recode(
      risk_type,
      !!!cancer_risk_labels
    )
  ) %>%
  arrange(
    risk_type,
    distance_ft,
    scenario
  )


# ------------------------------------------------------------
# 22. Save combined long-format CSV
# ------------------------------------------------------------

write.csv(
  cancer_distance_summary,
  file = cancer_distance_output_csv,
  row.names = FALSE
)


# ------------------------------------------------------------
# 23. Create wide reporting version
#
# The columns are grouped by scenario:
#
#   all Without-background statistics
#   all Background-only statistics
#   all Total statistics
# ------------------------------------------------------------

cancer_distance_summary_wide <-
  cancer_distance_summary %>%
  mutate(
    scenario = as.character(scenario),
    scenario = recode(
      scenario,
      "Without background" =
        "without_background",
      "Background only" =
        "background_only",
      "Total" =
        "total"
    )
  ) %>%
  pivot_wider(
    id_cols = c(
      risk_type,
      distance_ft,
      distance_m
    ),
    names_from = scenario,
    values_from = c(
      n,
      min,
      median,
      mean,
      max,
      sd,
      se
    ),
    names_glue = "{scenario}_{.value}"
  ) %>%
  select(
    risk_type,
    distance_ft,
    distance_m,

    # Without background
    without_background_n,
    without_background_min,
    without_background_median,
    without_background_mean,
    without_background_max,
    without_background_sd,
    without_background_se,

    # Background only
    background_only_n,
    background_only_min,
    background_only_median,
    background_only_mean,
    background_only_max,
    background_only_sd,
    background_only_se,

    # Total / with background
    total_n,
    total_min,
    total_median,
    total_mean,
    total_max,
    total_sd,
    total_se
  ) %>%
  arrange(
    risk_type,
    distance_ft
  )

cancer_distance_summary_wide


# ------------------------------------------------------------
# 24. Save Excel workbook
# ------------------------------------------------------------

cancer_distance_wb <- createWorkbook()


# Long-format table
addWorksheet(
  cancer_distance_wb,
  "Long summary"
)

writeData(
  cancer_distance_wb,
  sheet = "Long summary",
  x = cancer_distance_summary
)


# Wide-format table
addWorksheet(
  cancer_distance_wb,
  "Wide summary"
)

writeData(
  cancer_distance_wb,
  sheet = "Wide summary",
  x = cancer_distance_summary_wide
)


# Spatial grid counts
addWorksheet(
  cancer_distance_wb,
  "Distance grid counts"
)

writeData(
  cancer_distance_wb,
  sheet = "Distance grid counts",
  x = distance_grid_counts
)


# Background-only minimum-value check
addWorksheet(
  cancer_distance_wb,
  "Background value check"
)

writeData(
  cancer_distance_wb,
  sheet = "Background value check",
  x = background_min_by_chemical
)


# ------------------------------------------------------------
# 25. Format Excel workbook
# ------------------------------------------------------------

freezePane(
  cancer_distance_wb,
  sheet = "Long summary",
  firstRow = TRUE
)

freezePane(
  cancer_distance_wb,
  sheet = "Wide summary",
  firstRow = TRUE
)

freezePane(
  cancer_distance_wb,
  sheet = "Distance grid counts",
  firstRow = TRUE
)

freezePane(
  cancer_distance_wb,
  sheet = "Background value check",
  firstRow = TRUE
)

setColWidths(
  cancer_distance_wb,
  sheet = "Long summary",
  cols = 1:ncol(cancer_distance_summary),
  widths = "auto"
)

setColWidths(
  cancer_distance_wb,
  sheet = "Wide summary",
  cols = 1:ncol(cancer_distance_summary_wide),
  widths = "auto"
)

setColWidths(
  cancer_distance_wb,
  sheet = "Distance grid counts",
  cols = 1:ncol(distance_grid_counts),
  widths = "auto"
)

setColWidths(
  cancer_distance_wb,
  sheet = "Background value check",
  cols = 1:ncol(background_min_by_chemical),
  widths = "auto"
)


# ------------------------------------------------------------
# 26. Save workbook
# ------------------------------------------------------------

saveWorkbook(
  cancer_distance_wb,
  file = cancer_distance_output_xlsx,
  overwrite = TRUE
)
```


























